### 2024-06-04 manuscript.Rmd
manuscript update

---
title: "Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients"
subtitle: A single-center registry-based retrospective cohort
output: html_document
author: "Josef Al-Khalili"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abstract
========
*Introduction:* Trauma is responsible for 9% of global deaths annually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes using
Audit Filters (AFs) and interdisciplinary Morbidity and Mortality
Conferences (MMC). AFs are based on criteria for standard care and can
function as a screening tool of eligible patient cases for discussion at
MMCs. One of the goals with MMCs is to determine potential Opportunities
for Improvement (OFI). *Aims:* This study aimed to evaluate the
performance of individual AFs in predicting OFIs and to assess their
relevance as a screening tool for Morbidity and Mortality Conferences
(MMC). *Material and Methods:* Register data from 8,309 patient cases
were collected for analysis. The sensitivity, specificity and Area Under
the Receiver Operating Characteristics Curve (AUC) was calculated to
assess performance. An empirical bootstrapping test of AUC values was
conducted to compare the performance of each AF to coincidence (AUC =
0.5) and minimum AUC for diagnostic tests (AUC = 0.8). *Results:* Four
out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
AFs demonstrated a correlation with OFI, with AUC ranging from 0.51 to
0.63. The AF "\> 60 min until first intervention" exhibited the highest
AUC (0.63, 95% CI [0.61-0.65]). *Conclusion:* No individual AF seemed to
perform well in predicting OFI, which suggest rethinking the current
screening method for MMCs.

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============
Trauma, defined as a physical injury and
the body's associated response, accounts for 8-9%
of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed road traffic injuries [@vos2020global;
@roth2018global; @SweTrau2022Arsrapport]. Furthermore, trauma is responsible for the top cause for disability-adjusted life-years
(DALY) [@vos2020global] and associated with 36% not returning to work within one year
[@david2022measuring] and a higher
propensity towards depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course]. Trauma is
associated with . Attempts have been made to estimate the global
cost of trauma, which is a complicated task due to its variability.
However, it is estimated that the cost of road traffic injuries alone
accounts for 2% of gross domestic product (GDP) in high-income countries
[@world2014injuries].

## Trauma Care Systems

In order to reduce impact trauma care systems are in place. 
Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals
[@American2022resources]. Trauma care is divided into three components:
pre-hospital, in-hospital and post-hospital. The pre-hospital component
constitutes first responders, basic life support and advanced life
support [@world2005prehospital]. The in-hospital component is defined as
the moment a trauma patient is admitted to a health care facility, often
a trauma centre. The post-hospital component involves the rehabilitation
and recovery of treated trauma patients.

Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. Despite this, even established trauma care
systems are prone to errors due to the complex nature of trauma and its
susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities]. Studies between 1990 and 2014 have shown
a pooled preventable death rate of 20%, with more recent studies showing
a lower preventable death rate [@kwon2014pooled]. For example, a recent
study in a Swedish Level I trauma centre found a preventable death rate
of 4% [@ghorbani2018analysis].

## Quality Improvement Programmes

To minimise errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world
[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
facilitate discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
commonly used: a multidisciplinary mortality and morbidity conference
(MMC), a preventable death review panel, and the use of AFs also known
as quality indicators [@world2009guidelines].

### Morbidity and Mortality Conferences

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, improvement proposals
and implementing such [@gregor2016morbidity].

The conference participants are healthcare professionals involved in
trauma care and its processes, but not necessarily the specific patient
case. The selection of patient cases varies, but common themes for
discussion are unexpected mortality, unexpected morbidity and errors
[@orlander2003morbidity].

Implemented effectively, MMCs serve as an important tool for error
reduction without being perceived as negative feedback for individual
care providers, but rather as an incentive for reporting. These
conferences promote a more comprehensive, systems-based approach that
emphasises task analysis, teamwork and quality improvement
[@lazzara2022morbidity]. The introduction of MMC has been shown to
increase morbidity and mortality reporting [@kong2016analysis].

### Audit Filters and Opportunities for Improvement

AFs, also known as quality indicators are established criteria based on
standardised care. Any deviation from the AFs should raise the attention
of health care professionals with the purpose of improving the quality
of care [@American2022resources]. An AF can involve outcomes such as
"death" or criteria related to guidelines and gold standards such as
"the placement of two large bore intravenous lines within 15 minutes
from arrival to a health care facility" [@world2009guidelines].

Previous studies have shown a reduction in trauma-related mortality
after the implementation of AFs [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Per contra, newer studies seem to differ. In
2009, an extensive review by Evans et al. [@evans2009audit] sought to
determine if AFs could be used in improving processes in trauma care.
They found no studies successfully providing evidence for the use of
AFs. More than 1,500 AFs have been identified in a review by Stelfox et
al. [@stelfox2010quality]. Notably, they did not find a common set of
clearly defined, evidence-based, broadly accepted AFs for evaluating the
quality of trauma care. The perceived usefulness of audit filters still
remains high [@berg2022perceived]

Opportunities for improvement (OFIs) are the endpoint of patient care.
Following the review of individual patient cases there is a decision
made regarding the existence of an OFI [@American2022resources]. It has
been shown that such review processes are associated with high-quality
trauma care [@santana2014development]. OFIs are typically associated
with failures in initial care [@vioque2014classifying], specifically in
airway management, fluid resuscitation, haemorrhage control and chest
injury management [@sanddal2011analysis; @roy2017learning;
@o2013opportunities].

There is lacking evidence for the use of AFs for improving trauma care
[@evans2009audit; @stelfox2010quality] and the focus has been on
mortality as outcome measure [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Furthermore, the use of AFs has also been
associated with high frequencies of false positive OFI predictions,
ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
some AFs do not appear to correlate with OFIs at all [@lewis2017delay].

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2024-06-08 manuscript.Rmd
update

@@
 the body's associated response, accounts for 8-9%
-of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed road traffic injuries [@vos2020global;
-@roth2018global; @SweTrau2022Arsrapport]. Furthermore, trauma is responsible for the top cause for disability-adjusted life-years
-(DALY) [@vos2020global] and associated with 36% not returning to work within one year
-[@david2022measuring] and a higher
-propensity towards depression, post-traumatic stress disorder (PTSD) and
-negative effects on cognitive function [@kessler1995posttraumatic;
-@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course]. Trauma is
-associated with . Attempts have been made to estimate the global
-cost of trauma, which is a complicated task due to its variability.
-However, it is estimated that the cost of road traffic injuries alone
+of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed to mainly road traffic injuries [@vos2020global;
+@roth2018global; @SweTrau2022Arsrapport].
+Furthermore, trauma is responsible for the top cause for disability-adjusted life-years (DALY) [@vos2020global] and a higher propensity towards mental disorders and cognitive decline [@kessler1995posttraumatic;
+@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course].
+This negatively affects individuals, communities and society as a whole [@newnam2014impacts].It is estimated that the cost of road traffic injuries alone
 accounts for 2% of gross domestic product (GDP) in high-income countries
@@
 
-## Trauma Care Systems
+To reduce trauma burden, quality improvement (QI) programs
+have been established by the World Health Organization (WHO) and the
+International Association for Trauma Surgery and Intensive Care (IATSIC)
+[@world2009guidelines]. These programs form the core of current trauma
+care systems and their development around the world
+[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
+facilitate discussion and analysis of the trauma process in order to
+take corrective action for future success. The following techniques are
+commonly used: a multidisciplinary mortality and morbidity conference
+(MMC), a preventable death review panel, and the use of AFs also known
+as quality indicators [@world2009guidelines].
 
-In order to reduce impact trauma care systems are in place. 
 Trauma care is a multidisciplinary, complex and time-critical health
-care service provided by hospitals
-[@American2022resources]. Trauma care is divided into three components:
-pre-hospital, in-hospital and post-hospital. The pre-hospital component
-constitutes first responders, basic life support and advanced life
-support [@world2005prehospital]. The in-hospital component is defined as
-the moment a trauma patient is admitted to a health care facility, often
-a trauma centre. The post-hospital component involves the rehabilitation
-and recovery of treated trauma patients.
-
+care service provided by hospitals, often trauma centres.
+[@American2022resources].
 Trauma centres have a significantly lower mortality rate compared to
@@
 of 4% [@ghorbani2018analysis].
-
-## Quality Improvement Programmes
-
-To minimise errors in trauma care, quality improvement (QI) programs
-have been established by the World Health Organization (WHO) and the
-International Association for Trauma Surgery and Intensive Care (IATSIC)
-[@world2009guidelines]. These programs form the core of current trauma
-care systems and their development around the world
-[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
-facilitate discussion and analysis of the trauma process in order to
-take corrective action for future success. The following techniques are
-commonly used: a multidisciplinary mortality and morbidity conference
-(MMC), a preventable death review panel, and the use of AFs also known
-as quality indicators [@world2009guidelines].
 
@@
 
+##Participants
+After excluding patients
+under 15 years old and patients not screened for OFIs, a total of 8,309
+individuals were included in the study. 7,797 out of 8,309 (93.8%)
+patient cases did not have an OFI and 512 out of 8,309 (6.2%) had an
+OFI. Figure 1 presents a flowchart showcasing the exclusion and inclusion of
+the patient cases from the trauma registry.
+
+Most trauma patients were men (n = 5,755, 69%). Patient characteristics
+varied depending on whether patients had an OFI or not. Respectively,
+the mean age was significantly higher\
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
+`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`
+and
+the ISS was significantly higher
+`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
+`r inline_text(tableOne, variable = ISS, column = stat_1)`,
+`r inline_text(tableOne, variable = ISS, column = "p.value")`. The
+number of intubations in the emergency department was significantly
+higher
+`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
+vs
+`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`,
+`r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
+Patients with OFI had longer times to definitive treatment from hospital
+arrival compared to patients without OFI
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
+vs
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`,
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
+The time to the first CT was also longer in patients with OFIs compared
+to patients without
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`,
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`.
+
+The number of OFIs were highest in patients admitted to the intensive
+care unit
+`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)`
+and surgical ward
+`r inline_text(tableOne, variable = host_care_level, level = "Surgical ward", column = stat_2)`.
+
+Gender `r inline_text(tableOne, variable = Gender, column = "p.value")`,
+the mean systolic blood pressure
+`r inline_text(tableOne, variable = ed_sbp_value, column = "p.value")`,
+death at 30 days
+`r inline_text(tableOne, variable = res_survival, column = "p.value")`
+and trauma team activation
+`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was not
+significantly correlated with OFI.
+
+Notably, there was a statistical significant correlation between GCS at
+ED and OFI (p \< 0.001).
+
+Table 2 shows the missing data for each AF. As mentioned in the method
+section, these values were interpreted as "not flagged" with an AF. The
+variables with the highest amount of missing data were "time to
+definitive treatment" (n = 5,990, 72.1%) and "time to first CT" (n =
+1,012, 12.2%). The smallest amount of missing data was found in
+"SBP\<90" (n = 13, 0.2%).
+
+## Performance of Individual Audit Filters
+
+The performance of each AF was determined through a calculation of
+sensitivity and specificity which is demonstrated in Table 4.
+
+The number of patient cases flagged by each individual AF varied between
+107 for "GCS\<9 and not intubated" and 3907 for "\>30 min until CT". The
+AF with the highest sensitivity was "\>30 min until CT" (59.2%, 95% CI
+[55.2-63.8]). The lowest sensitivity was seen in "GCS\<9 and not
+intubated" (1.2%, 95% CI [0.2-2]). The AF with the highest specificity
+was "GCS\<9 and not intubated" (98.7%, 95% CI [98.5-98.9]). The lowest
+specificity was seen in "\>30 min until CT" (53.8%, 95% CI [52.7-54.8]).
+Notably, the confidence intervals varied between sensitivity and
+specificity, with sensitivity displaying a greater interval.
+
+Table 5 summarises the performance of each individual AF in identifying
+OFIs in trauma care by AUC. Four of the 11 AFs, "SBP\<90", "dead at 30
+days", "GCS\<9 and not intubated" and "CPR and thoracotomy" showed a
+performance in predicting OFIs equal to that of coincidence (AUC = 0.5).
+The highest AUC was seen in "\>60 min until first intervention" (0.63,
+95% CI [0.61-0.65]). All AFs showed significantly lower AUC compared to
+the minimum standard for diagnostic tests (AUC = 0.8) (p \< 0.001).
+
+Table 2: Comparison of missing values for each AF. Missing values is
+shown as a number and percentage of the total number of patient cases
+included in the study (n = 8309).\
+`r tableTwo` Table 3: Patient characteristics of individuals with and
+without an OFI. Missing values are shown only when they exist.
+`r tableOne` Table 4: Comparison of performance for each AF in
+predicting OFIs. Performance is defined by sensitivity (%) and
+specificity (%). A 95% confidence interval is shown together with each
+performance indicator. The number of cases flagged by each filter is
+displayed.\
+`r tableThree` Table 5: Comparison of performance for each AF in
+predicting OFIs. Performance is defined by the area under the receiver
+operating characteristics curve. P-values show the significance of AUC
+compared to coincidence (AUC = 0.5) and to the minimum standard for
+diagnostic tests (AUC = 0.8) [@nahm2022receiver]. The p-values were
+calculated using an empirical bootstrap test. Furthermore, a 95%
+confidence interval was used. `r tableFour`
+
+```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
+knitr::include_graphics("ofi-flowchart.pdf")
+```
+
+Figure 1: Flowchart of the patient selection process. Shows the patients
+excluded and included in this study.
 <!--
@@
 ==========
+This study showed that AFs do not perform well in predicting OFIs.
+No single AF presented with a performance of value in predicting OFI in trauma care patients.
+Some AF showed a correlation with OFI but not enough to be considered useful.
 
+Patient characteristics are similar to previous studier showing twofold prevalence of male trauma patients compared to females [@roth2018global;@SweTrau2022Arsrapport]. Previous studies have shown
+
+A previous study conducted at Karolinska University Hospital by Albaaj
+et al. [@albaaj2023patient] sought to identify patient factors
+associated with OFIs and found that a higher ISS, longer times to CT,
+longer times to first intervention and higher levels of care was
+correlated with OFIs. These findings correspond with our analysis of
+patient characteristics and AFs linked to these factors.
+
+Higher ISS and higher levels of care in patients with OFIs as opposed to
+patients without OFIs could be explained by the greater complexity of
+such trauma cases. Greater complexity requires more interventions which
+could allow for more mistakes [@albaaj2023patient].
+
+Notably, GCS at ED showed a correlation with OFI even though the median
+GCS and its IQR was identical for individuals with OFI and without. Upon
+further analysis this could be explained by the difference in
+distribution outside the IQR.
+
+AFs generally exhibited a high specificity relative to their sensitivity
+which may be attributed to the outcome being rather uncommon, whilst
+every individual AF misses the majority of OFIs. This could suggest that
+the AFs are too rigid in their criteria. On the contrary, a too
+inclusive criteria might flag several patients, who do not necessarily
+have an OFI [@willis2008interpreting].
+
+In 2008, Willis et al. [@willis2008interpreting] assessed 14 audit
+filters, finding only three to be associated with an increased risk of
+poor outcomes. One AF, "GCS\<9 and not intubated", also used in this
+study, similarly showed no significant ability to predict OFIs. An
+explanation for this could be that an abnormal GCS as a result from
+intoxication is not comparable to an abnormal GCS due to head injury for
+example, suggesting that the AF might be more reflective of clinical
+decision rather than quality of care [@willis2008interpreting]. However,
+according to Sanddal et al. [@sanddal2011analysis] airway management is
+associated with OFI suggesting that an alternative audit filter related
+to airway management might potentially perform differently.
+
+Both AFs measuring delay: "\>60 until first intervention" and "\>30min
+until first CT", showed a correlation with OFI. This is comparable to
+what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
+correlating to preventable death. Furthermore, Ghorbani et al.
+[@ghorbani2018analysis] identified delay in treatment, delay to CT, and
+clinical judgement errors as common errors in trauma care. A more recent
+study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
+significant reduction in mortality when patients received a whole-body
+CT within 10 minutes of arrival at the ED. It is important to note that
+OFI in this study was not only a reflection of mortality but a broader
+concept encompassing different aspects of care quality. If anything, it
+highlights the difference between hospitals, which AFs are in use, and
+how they are being assessed.
+
+AFs that present with a high risk of inevitable death showed poor"Dead
+at 30 days" and "CPR and thoracotomy", could be explained by inevitable
+mortality, thus presenting with a smaller time frame for mistakes to be
+made.
+
+##Interpretation
+
+AFs are static in their criteria, contrary to health care in general.
+Consequently, the utility of AFs may diminish if confronted with
+emerging challenges or upon resolution of the quality gaps they were
+designed for. The static nature of AFs highlight their need for
+consistent and evidence-based review to determine current usefulness
+[@santana2014development]. However, it is worth pointing out that the
+review process might present with costs greater than the potential
+benefit from AFs.
+
+AFs face noteworthy challenges due to over-reliance, bias and
+confounders. Firstly, while AFs may help in determining the existence of
+OFI, they could overshadow other factors that may contribute to that
+analysis, leading to over-reliance. Secondly, devising AFs comes with an
+unintended bias. They might disproportionally affect certain groups of
+people failing to recognise patient cases that fall outside the scope of
+AF criteria [@willis2008interpreting]. For example, AFs primarily focus
+on aspects of trauma care mostly relevant to medical doctors despite the
+interdisciplinary nature of trauma care. This is especially
+counterintuitive given the broad definition of OFIs. Lastly, confounding
+variables adds to the uncertainty of AFs performance. For example,
+patients receiving the same interventions at different time points may
+do so as a function of injury severity not quality of care
+[@willis2008interpreting].
+
+## Limitations
+
+Opportunity for improvement, while dichotomously defined, includes a
+diverse set of outcomes. This convolutes the process of creating AFs
+that perform well in predicting OFIs. The current screening system for
+OFIs might also introduce bias since the AFs would favour the
+identification of some, but not all, errors.
+
+Each AF was individually related to OFI. Consequently, even if an AF
+demonstrated a true positive correlation with an OFI, instances where
+the same patient case had multiple AFs introduce a degree of uncertainty
+regarding which specific AF accurately predicted the outcome.
+
+The number of missing values varied greatly among the AFs. As stated
+above, missing values from each AF were coded as "not flagged" by an AF.
+This means that some instances where patient cases were not flagged,
+might actually meet the AF criteria, but for some reason not documented.
+This reduces the validity of some AFs that featured high amounts of
+missing data.
+
+## Generalizability
+
+AFs are widely used but generalizability is complicated due to the lack
+of common grounds on what they are, what purpose they serve and how they
+shall be used [@stelfox2010quality; @stelfox2011evidence;
+@berg2022perceived].
+
+Our findings suggest that AFs perform poorly in predicting OFIs. The AFs
+with performance exceeding coincidence (AUC = 0.5) might be a beneficial
+addition to an already existing quality improvement programme or could
+be used to fine-tune AFs already in use to plug hospitals' quality gaps.
+Due to the study being conducted in a single, trauma level I centre in
+Stockholm it is unrealistic to assume applicability of these findings at
+other hospitals, in other countries, facing unique challenges
+[@world2009guidelines]. For example, Low and middle income countries
+seem to benefit from certain AFs that high income countries do not and
+vice versa [@berg2022perceived].
+
+## Strengths
+
+As stated in the method section, some AFs were manually created from
+registry data instead of relying on the selection process by the
+specialised nurse (see table 1). This, in order to minimise
+misclassification bias.
+
+Furthermore, there has been very limited high quality evidence
+supporting the efficacy of AFs in improving trauma care
+[@evans2009audit; @stelfox2010quality] and predicting OFI
+[@sanddal2011analysis; @roy2017learning]. This study shows comparable
+results. Previous studies also focus primarily on AFs' ability to reduce
+mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In
+this study the morbidity of patient cases is also taken into account
+using OFIs as an umbrella term that includes factors ranging from bad
+documentation to preventable death.
+
+From 2017 all trauma patients admitted to Karolinska University Hospital
+were included in the registry reducing the risk of bias in inclusion of
+patient cases.
+
+## Future Studies
+
+With the ongoing academic, political, and public attention towards
+patient safety and quality of healthcare [@evans2009audit], it is
+important to discover optimal methods for monitoring and enhancing the
+care provided within trauma systems.
+
+Given the expanding field of artificial intelligence and machine
+learning, newer studies should focus on using these tools to screen for
+patient cases at risk of OFIs. This will diminish misclassification
+rates and ensure consistent and sustained high performance, attributable
+to an algorithm's capability to adapt and learn from the prevailing
+context of trauma care. This approach is contemporary but not
+unexplored. A preprint is currently available by Attergrim et al.
+[@attergrim2023predicting] exploring the idea of machine learning models
+as a screening tool for potential OFIs. They found that machine learning
+models outperformed the AFs currently used at Karolinska University
+Hospital in predicting OFIs.
+
+Should the use of AFs remain, it is important to ensure its performance
+by routinely updating them in order to better reflect the current
+quality gaps.
 <!--
@@
 
-<!-- Give a short conclusion, corresponding to the aim -->
+AFs are widely used as a tool in quality improvement programmes and
+should reflect the current gap in trauma care quality. Strong evidence
+supporting their efficacy in quality improvement is scarce. In this
+study, no audit filter seemed to show promising performance in
+predicting OFIs with AUC ranging from 0.5 to 0.63. These results
+highlight the importance of continually evaluating the relevance of AFs.
+However, this process takes time and resources and using an alternative
+to AFs is recommended.
 
@@
 ==========
-
-<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2024-06-08 manuscript.Rmd
update

@@
 This study showed that AFs do not perform well in predicting OFIs. 
-No single AF presented with a performance of value in predicting OFI in trauma care patients.
-Some AF showed a correlation with OFI but not enough to be considered useful.
-
-Patient characteristics are similar to previous studier showing twofold prevalence of male trauma patients compared to females [@roth2018global;@SweTrau2022Arsrapport]. Previous studies have shown
-
-A previous study conducted at Karolinska University Hospital by Albaaj
-et al. [@albaaj2023patient] sought to identify patient factors
-associated with OFIs and found that a higher ISS, longer times to CT,
-longer times to first intervention and higher levels of care was
-correlated with OFIs. These findings correspond with our analysis of
-patient characteristics and AFs linked to these factors.
-
-Higher ISS and higher levels of care in patients with OFIs as opposed to
-patients without OFIs could be explained by the greater complexity of
-such trauma cases. Greater complexity requires more interventions which
-could allow for more mistakes [@albaaj2023patient].
-
-Notably, GCS at ED showed a correlation with OFI even though the median
-GCS and its IQR was identical for individuals with OFI and without. Upon
-further analysis this could be explained by the difference in
-distribution outside the IQR.
-
 AFs generally exhibited a high specificity relative to their sensitivity
@@
 
-In 2008, Willis et al. [@willis2008interpreting] assessed 14 audit
-filters, finding only three to be associated with an increased risk of
-poor outcomes. One AF, "GCS\<9 and not intubated", also used in this
-study, similarly showed no significant ability to predict OFIs. An
-explanation for this could be that an abnormal GCS as a result from
-intoxication is not comparable to an abnormal GCS due to head injury for
-example, suggesting that the AF might be more reflective of clinical
-decision rather than quality of care [@willis2008interpreting]. However,
-according to Sanddal et al. [@sanddal2011analysis] airway management is
-associated with OFI suggesting that an alternative audit filter related
-to airway management might potentially perform differently.
-
-Both AFs measuring delay: "\>60 until first intervention" and "\>30min
-until first CT", showed a correlation with OFI. This is comparable to
-what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
-correlating to preventable death. Furthermore, Ghorbani et al.
-[@ghorbani2018analysis] identified delay in treatment, delay to CT, and
-clinical judgement errors as common errors in trauma care. A more recent
-study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
-significant reduction in mortality when patients received a whole-body
-CT within 10 minutes of arrival at the ED. It is important to note that
-OFI in this study was not only a reflection of mortality but a broader
-concept encompassing different aspects of care quality. If anything, it
-highlights the difference between hospitals, which AFs are in use, and
-how they are being assessed.
-
-AFs that present with a high risk of inevitable death showed poor"Dead
-at 30 days" and "CPR and thoracotomy", could be explained by inevitable
-mortality, thus presenting with a smaller time frame for mistakes to be
-made.
-
-##Interpretation
-
 AFs are static in their criteria, contrary to health care in general.
@@
 emerging challenges or upon resolution of the quality gaps they were
-designed for. The static nature of AFs highlight their need for
+designed for. This highlights their need for
 consistent and evidence-based review to determine current usefulness
-[@santana2014development]. However, it is worth pointing out that the
+[@santana2014development]. However, such
 review process might present with costs greater than the potential
@@
 
-AFs face noteworthy challenges due to over-reliance, bias and
-confounders. Firstly, while AFs may help in determining the existence of
-OFI, they could overshadow other factors that may contribute to that
-analysis, leading to over-reliance. Secondly, devising AFs comes with an
-unintended bias. They might disproportionally affect certain groups of
+While AFs may help determining
+OFIs, they could overshadow other factors contributing to that
+analysis, resulting in over-reliance. AFs might disproportionally affect certain groups of
 people failing to recognise patient cases that fall outside the scope of
@@
 
+Higher ISS, longer times to CT, longer times to first intervention and higher levels of care were correlated with OFIs, corresponding with a previous study conducted at Karolinska University Hospital by Albaaj
+et al.
+[@albaaj2023patient].
+Greater severity and higher levels of care call for greater complexity, requiring more interventions, allowing for more mistakes.
+This could explain their correlation with OFI[@albaaj2023patient].
+
+Notably, GCS at ED showed a correlation with OFI even though the median
+GCS and its IQR was identical for individuals with OFI and without. Upon
+further analysis this could be explained by the difference in
+distribution outside the IQR.
+
+The AF "GCS<9 and not intubated" showed no significant ability to predict OFIs, similarly shown by Willis et al. [@willis2008interpreting] in 2008. An
+explanation for this could be that an abnormal GCS as a result from
+intoxication versus head injury for example are not comparable, suggesting that the AF might be more reflective of clinical
+decision rather than quality of care [@willis2008interpreting]. However, according to Sanddal et al. [@sanddal2011analysis] airway management was
+associated with OFI, suggesting that an alternative AF related
+to airway management might perform differently.
+
+Both AFs measuring delay: ">60 until first intervention" and ">30min
+until first CT", showed a correlation with OFI. This is comparable to
+what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
+correlating to preventable death. Furthermore, Ghorbani et al.
+[@ghorbani2018analysis] identified delay in treatment and delay to CT as common errors in trauma care. A more recent
+study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
+significant reduction in mortality when patients received a whole-body
+CT within 10 minutes of arrival at the ED. Notably,
+OFI in this study was not soley a reflection of mortality but also morbidity. Rather, it
+highlights the difference between hospitals, which AFs are in use, and
+how they are assessed.
+
+AFs that present with a high risk of inevitable death: "Dead
+at 30 days" and "CPR and thoracotomy" showed no correlation with OFI. This could be explained by inevitable
+mortality, thus presenting with a smaller time frame for mistakes to be
+made.
+
+##Interpretation
+
 ## Limitations
 
-Opportunity for improvement, while dichotomously defined, includes a
+OFI, while dichotomously defined, includes a
 diverse set of outcomes. This convolutes the process of creating AFs
-that perform well in predicting OFIs. The current screening system for
-OFIs might also introduce bias since the AFs would favour the
+that perform well in predicting OFIs. AFs would favour the
 identification of some, but not all, errors.
 
-Each AF was individually related to OFI. Consequently, even if an AF
+Each AF was individually related to OFI. Consequently, if an AF
 demonstrated a true positive correlation with an OFI, instances where
-the same patient case had multiple AFs introduce a degree of uncertainty
-regarding which specific AF accurately predicted the outcome.
+the same patient case had multiple AFs introduce uncertainty
+regarding which specific AF truthfully predicted the OFI.
 
@@
 
-Our findings suggest that AFs perform poorly in predicting OFIs. The AFs
+Our findings suggest that AFs perform poorly in predicting an OFI. The AFs
 with performance exceeding coincidence (AUC = 0.5) might be a beneficial
-addition to an already existing quality improvement programme or could
+addition to an already existing QI programme or
 be used to fine-tune AFs already in use to plug hospitals' quality gaps.
@@
 Stockholm it is unrealistic to assume applicability of these findings at
-other hospitals, in other countries, facing unique challenges
+other hospitals facing unique challenges
 [@world2009guidelines]. For example, Low and middle income countries

### 2024-06-10 manuscript.Rmd
update

@@
 csl: vancouver.csl
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Abstract
-========
-*Introduction:* Trauma is responsible for 9% of global deaths annually
+# Abstract
+
+## Background
+
+Trauma is responsible for 9% of global deaths annually
 and has a significant impact on individuals and society. Quality
@@
 MMCs. One of the goals with MMCs is to determine potential Opportunities
-for Improvement (OFI). *Aims:* This study aimed to evaluate the
+for Improvement (OFI).This study aimed to evaluate the
 performance of individual AFs in predicting OFIs and to assess their
 relevance as a screening tool for Morbidity and Mortality Conferences
-(MMC). *Material and Methods:* Register data from 8,309 patient cases
+(MMC).
+
+## Methods
+
+Register data from 8,309 patient cases
 were collected for analysis. The sensitivity, specificity and Area Under
@@
 conducted to compare the performance of each AF to coincidence (AUC =
-0.5) and minimum AUC for diagnostic tests (AUC = 0.8). *Results:* Four
+0.5) and minimum AUC for diagnostic tests (AUC = 0.8).
+
+## Results
+
+Four
 out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
@@
 0.63. The AF "\> 60 min until first intervention" exhibited the highest
-AUC (0.63, 95% CI [0.61-0.65]). *Conclusion:* No individual AF seemed to
+AUC (0.63, 95% CI [0.61-0.65]).
+
+## Conclusion
+
+No individual AF seemed to
 perform well in predicting OFI, which suggest rethinking the current
@@
 
-Background
-----------
+# Introduction
 
-Methods
--------
+Trauma, defined as a physical injury and the body's associated response,
+accounts for 8-9% of global deaths annually [@world2014injuries;
+@roth2018global], mainly affecting younger individuals in low- and
+middle-income countries attributed to mainly road traffic injuries
+[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Furthermore,
+trauma is responsible for the top cause for disability-adjusted
+life-years (DALY) [@vos2020global] and a higher propensity towards
+mental disorders and cognitive decline [@kessler1995posttraumatic;
+@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
+@visser2017course]. This negatively affects individuals, communities and
+society as a whole [@newnam2014impacts].It is estimated that the cost of
+road traffic injuries alone accounts for 2% of gross domestic product
+(GDP) in high-income countries [@world2014injuries].
 
-Results
--------
-
-Conclusion
-----------
-
-Introduction
-============
-Trauma, defined as a physical injury and
-the body's associated response, accounts for 8-9%
-of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed to mainly road traffic injuries [@vos2020global;
-@roth2018global; @SweTrau2022Arsrapport].
-Furthermore, trauma is responsible for the top cause for disability-adjusted life-years (DALY) [@vos2020global] and a higher propensity towards mental disorders and cognitive decline [@kessler1995posttraumatic;
-@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course].
-This negatively affects individuals, communities and society as a whole [@newnam2014impacts].It is estimated that the cost of road traffic injuries alone
-accounts for 2% of gross domestic product (GDP) in high-income countries
-[@world2014injuries].
-
-To reduce trauma burden, quality improvement (QI) programs
-have been established by the World Health Organization (WHO) and the
-International Association for Trauma Surgery and Intensive Care (IATSIC)
+To reduce trauma burden, quality improvement (QI) programs have been
+established by the World Health Organization (WHO) and the International
+Association for Trauma Surgery and Intensive Care (IATSIC)
 [@world2009guidelines]. These programs form the core of current trauma
 care systems and their development around the world
-[@American2022resources]. QI programs consist of a series of recommended techniques aiming to
-facilitate discussion and analysis of the trauma process in order to
-take corrective action for future success. The following techniques are
-commonly used: a multidisciplinary mortality and morbidity conference
-(MMC), a preventable death review panel, and the use of AFs also known
-as quality indicators [@world2009guidelines].
+[@American2022resources]. QI programs consist of a series of recommended
+techniques aiming to facilitate discussion and analysis of the trauma
+process in order to take corrective action for future success. The
+following techniques are commonly used: a multidisciplinary mortality
+and morbidity conference (MMC), a preventable death review panel, and
+the use of AFs also known as quality indicators [@world2009guidelines].
 
@@
 care service provided by hospitals, often trauma centres.
-[@American2022resources].
-Trauma centres have a significantly lower mortality rate compared to
-non-trauma centres, and the maturity of a trauma centre correlates with
-lower mortality, demonstrating the importance of high-quality,
-specialised trauma care [@alharbi2021effectiveness; @mann1999systematic;
+[@American2022resources]. Trauma centres have a significantly lower
+mortality rate compared to non-trauma centres, and the maturity of a
+trauma centre correlates with lower mortality, demonstrating the
+importance of high-quality, specialised trauma care
+[@alharbi2021effectiveness; @mann1999systematic;
 @candefjord2020mortality]. Despite this, even established trauma care
@@
 susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
-@o2013opportunities]. Studies between 1990 and 2014 have shown
-a pooled preventable death rate of 20%, with more recent studies showing
-a lower preventable death rate [@kwon2014pooled]. For example, a recent
-study in a Swedish Level I trauma centre found a preventable death rate
-of 4% [@ghorbani2018analysis].
+@o2013opportunities]. Studies between 1990 and 2014 have shown a pooled
+preventable death rate of 20%, with more recent studies showing a lower
+preventable death rate [@kwon2014pooled]. For example, a recent study in
+a Swedish Level I trauma centre found a preventable death rate of 4%
+[@ghorbani2018analysis].
 
@@
 
+```{=html}
 <!-- Explain the scientific background and rationale for the
@@
 6-8 lines long. Support every statement with a citation -->
+```
+# Methods
 
-Methods 
-=======
-
-Study design
-------------
+## Study design
 
@@
 
-Setting
--------
+## Setting
 
+```{=html}
 <!-- Describe the setting, locations, and relevant dates, including
 periods of recruitment, exposure, follow-up, and data collection -->
+```
+## Participants
 
-Participants
-------------
-
+```{=html}
 <!-- Cohort study: Give the eligibility criteria, and the sources and
@@
 exposed and unexposed -->
-
+```
+```{=html}
 <!-- Case-control study: Give the eligibility criteria, and the
@@
 of controls per case -->
-
+```
+```{=html}
 <!-- Cross-sectional study: Give the eligibility criteria, and the
 sources and methods of selection of participants -->
+```
+## Variables and data sources/measurements
 
-Variables and data sources/measurements
----------------------------------------
-
+```{=html}
 <!-- Clearly define all outcomes, exposures, predictors, potential
@@
 of assessment methods if there is more than one group -->
-
-Bias 
-----
+```
+## Bias
 
@@
 
-Study size 
-----------
+## Study size
 
@@
 
-Quantitative variables
-----------------------
+## Quantitative variables
 
+```{=html}
 <!-- Explain how quantitative variables were handled in the
@@
 -->
+```
+## Statistical methods
 
-Statistical methods
--------------------
-
+```{=html}
 <!--
@@
 -->
+```
+# Results
 
-Results
-=======
-
-##Participants
-After excluding patients
-under 15 years old and patients not screened for OFIs, a total of 8,309
-individuals were included in the study. 7,797 out of 8,309 (93.8%)
-patient cases did not have an OFI and 512 out of 8,309 (6.2%) had an
-OFI. Figure 1 presents a flowchart showcasing the exclusion and inclusion of
-the patient cases from the trauma registry.
+##Participants After excluding patients under 15 years old and patients
+not screened for OFIs, a total of 8,309 individuals were included in the
+study. 7,797 out of 8,309 (93.8%) patient cases did not have an OFI and
+512 out of 8,309 (6.2%) had an OFI. Figure 1 presents a flowchart
+showcasing the exclusion and inclusion of the patient cases from the
+trauma registry.
 
@@
 `r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
-`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`
-and
+`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")` and
 the ISS was significantly higher
@@
 and trauma team activation
-`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was not
-significantly correlated with OFI.
+`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was
+not significantly correlated with OFI.
 
@@
 Figure 1: Flowchart of the patient selection process. Shows the patients
-excluded and included in this study.
-<!--
+excluded and included in this study. <!--
 
@@
 
-You can also mix text and code, so called inline code, like this: `r 2+5`.
+You can also mix text and code, so called inline code, like this:
+`r 2+5`.
 
-Discussion
-==========
-This study showed that AFs do not perform well in predicting OFIs. 
-AFs generally exhibited a high specificity relative to their sensitivity
-which may be attributed to the outcome being rather uncommon, whilst
-every individual AF misses the majority of OFIs. This could suggest that
-the AFs are too rigid in their criteria. On the contrary, a too
-inclusive criteria might flag several patients, who do not necessarily
-have an OFI [@willis2008interpreting].
+# Discussion
 
-AFs are static in their criteria, contrary to health care in general.
-Consequently, the utility of AFs may diminish if confronted with
-emerging challenges or upon resolution of the quality gaps they were
-designed for. This highlights their need for
-consistent and evidence-based review to determine current usefulness
-[@santana2014development]. However, such
-review process might present with costs greater than the potential
-benefit from AFs.
-
-While AFs may help determining
-OFIs, they could overshadow other factors contributing to that
-analysis, resulting in over-reliance. AFs might disproportionally affect certain groups of
-people failing to recognise patient cases that fall outside the scope of
-AF criteria [@willis2008interpreting]. For example, AFs primarily focus
-on aspects of trauma care mostly relevant to medical doctors despite the
-interdisciplinary nature of trauma care. This is especially
-counterintuitive given the broad definition of OFIs. Lastly, confounding
-variables adds to the uncertainty of AFs performance. For example,
-patients receiving the same interventions at different time points may
-do so as a function of injury severity not quality of care
-[@willis2008interpreting].
-
-Higher ISS, longer times to CT, longer times to first intervention and higher levels of care were correlated with OFIs, corresponding with a previous study conducted at Karolinska University Hospital by Albaaj
-et al.
-[@albaaj2023patient].
-Greater severity and higher levels of care call for greater complexity, requiring more interventions, allowing for more mistakes.
-This could explain their correlation with OFI[@albaaj2023patient].
-
-Notably, GCS at ED showed a correlation with OFI even though the median
-GCS and its IQR was identical for individuals with OFI and without. Upon
-further analysis this could be explained by the difference in
-distribution outside the IQR.
-
-The AF "GCS<9 and not intubated" showed no significant ability to predict OFIs, similarly shown by Willis et al. [@willis2008interpreting] in 2008. An
-explanation for this could be that an abnormal GCS as a result from
-intoxication versus head injury for example are not comparable, suggesting that the AF might be more reflective of clinical
-decision rather than quality of care [@willis2008interpreting]. However, according to Sanddal et al. [@sanddal2011analysis] airway management was
-associated with OFI, suggesting that an alternative AF related
-to airway management might perform differently.
-
-Both AFs measuring delay: ">60 until first intervention" and ">30min
-until first CT", showed a correlation with OFI. This is comparable to
-what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
-correlating to preventable death. Furthermore, Ghorbani et al.
-[@ghorbani2018analysis] identified delay in treatment and delay to CT as common errors in trauma care. A more recent
-study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
-significant reduction in mortality when patients received a whole-body
-CT within 10 minutes of arrival at the ED. Notably,
-OFI in this study was not soley a reflection of mortality but also morbidity. Rather, it
-highlights the difference between hospitals, which AFs are in use, and
-how they are assessed.
-
-AFs that present with a high risk of inevitable death: "Dead
-at 30 days" and "CPR and thoracotomy" showed no correlation with OFI. This could be explained by inevitable
-mortality, thus presenting with a smaller time frame for mistakes to be
-made.
-
-##Interpretation
+The aim of this study was to evaluate the performance of individual AFs in predicting OFIs and 
+to assess their suitability as a screening tool for MMCs.
+This study showed that no AFs was significantly successful in predicting OFIs.
+Four AFs ("SBP<90," "dead at 30 days," "GCS<9 and not intubated," and "CPR and thoracotomy") were equivalent to random chance in their predictive ability.
+The highest AUC was for ">60 minutes until first intervention,".
+Notably, all AFs performed significantly below the accepted standard for diagnostic tests (>0.8).
 
@@
 
-OFI, while dichotomously defined, includes a
-diverse set of outcomes. This convolutes the process of creating AFs
-that perform well in predicting OFIs. AFs would favour the
-identification of some, but not all, errors.
+OFI, while dichotomously defined, includes a diverse set of outcomes.
+This convolutes the process of creating AFs that perform well in
+predicting OFIs. AFs would favour the identification of some, but not
+all, errors.
 
@@
 demonstrated a true positive correlation with an OFI, instances where
-the same patient case had multiple AFs introduce uncertainty
-regarding which specific AF truthfully predicted the OFI.
+the same patient case had multiple AFs introduce uncertainty regarding
+which specific AF truthfully predicted the OFI.
 
@@
 
+##Interpretation
+
+Higher ISS, longer times to CT, longer times to first intervention and
+higher levels of care were correlated with OFIs, corresponding with a
+recent study conducted at Karolinska University Hospital by Albaaj et
+al. [@albaaj2023patient]. Notably, GCS at ED showed a correlation with OFI even though the median
+GCS and its IQR was identical for individuals with OFI and without. Upon
+further analysis this could be explained by the difference in
+distribution outside the IQR.
+
+The AF "GCS\<9 and not intubated" showed no significant ability to
+predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]
+in 2008. An explanation for this could be that an abnormal GCS as a
+result from intoxication versus head injury for example are not
+comparable, suggesting that the AF might be more reflective of clinical
+decision rather than quality of care [@willis2008interpreting]. However,
+according to Sanddal et al. [@sanddal2011analysis] airway management was
+associated with OFI, suggesting that an alternative AF related to airway
+management might perform differently.
+
+Both AFs measuring delay: "\>60 until first intervention" and "\>30min
+until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] found a correlation between delay
+and preventable death. Furthermore, Ghorbani et al.
+[@ghorbani2018analysis] identified delay in treatment and delay to CT as
+common errors in trauma care. A more recent study in Japan by Yamamoto
+et. al [@yamamoto2023immediate] showed a significant reduction in
+mortality when patients received a whole-body CT within 10 minutes of
+arrival at the ED. Notably, OFI in this study was not soley a reflection
+of mortality but also morbidity. Rather, it could highlight differences in quality gaps
+between hospitals.
+
+"Dead at 30 days"
+and "CPR and thoracotomy" showed no correlation with OFI. This could be
+explained by inevitable mortality, thus presenting with a smaller time
+frame for mistakes to be made.
+
+AFs
+generally exhibited a high specificity relative to their sensitivity
+which may be attributed to the outcome being rather uncommon, whilst
+every individual AF misses the majority of OFIs. This could suggest that
+the AFs are too rigid in their criteria. On the contrary, a too
+inclusive criteria might flag several patients, without an OFI [@willis2008interpreting].
+
 ## Generalizability
 
-AFs are widely used but generalizability is complicated due to the lack
-of common grounds on what they are, what purpose they serve and how they
+AFs, although widely used, there is lack
+of consensus on what they are, what purpose they serve and how they
 shall be used [@stelfox2010quality; @stelfox2011evidence;
-@berg2022perceived].
+@berg2022perceived]. Further, AFs are static in their criteria, contrary to health care in general.
+Consequently, the utility of AFs may diminish with emerging challenges
+or upon resolution of the quality gaps they were designed for. This
+highlights the need for consistent and evidence-based review to
+determine current usefulness [@santana2014development]. However, such
+review process might present with costs greater than the potential
+benefit from AFs.
 
-Our findings suggest that AFs perform poorly in predicting an OFI. The AFs
-with performance exceeding coincidence (AUC = 0.5) might be a beneficial
-addition to an already existing QI programme or
-be used to fine-tune AFs already in use to plug hospitals' quality gaps.
-Due to the study being conducted in a single, trauma level I centre in
-Stockholm it is unrealistic to assume applicability of these findings at
-other hospitals facing unique challenges
-[@world2009guidelines]. For example, Low and middle income countries
-seem to benefit from certain AFs that high income countries do not and
-vice versa [@berg2022perceived].
+While AFs may help determining OFIs, they could overshadow other factors
+contributing to that analysis, resulting in over-reliance. AFs might
+disproportionally affect certain groups of people failing to recognise
+patient cases that fall outside the scope of AF criteria
+[@willis2008interpreting]. For example, AFs primarily focus on aspects
+of trauma care mostly relevant to medical doctors despite the
+interdisciplinary nature of trauma care. This is especially
+counterintuitive given the broad definition of OFIs. Lastly, confounding
+variables adds to the uncertainty of AFs performance. For example,
+patients receiving the same interventions at different time points may
+do so as a function of injury severity not quality of care
+[@willis2008interpreting].
 
-## Strengths
-
-As stated in the method section, some AFs were manually created from
-registry data instead of relying on the selection process by the
-specialised nurse (see table 1). This, in order to minimise
-misclassification bias.
-
-Furthermore, there has been very limited high quality evidence
-supporting the efficacy of AFs in improving trauma care
-[@evans2009audit; @stelfox2010quality] and predicting OFI
-[@sanddal2011analysis; @roy2017learning]. This study shows comparable
-results. Previous studies also focus primarily on AFs' ability to reduce
-mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In
-this study the morbidity of patient cases is also taken into account
-using OFIs as an umbrella term that includes factors ranging from bad
-documentation to preventable death.
-
-From 2017 all trauma patients admitted to Karolinska University Hospital
-were included in the registry reducing the risk of bias in inclusion of
-patient cases.
+Due to the
+study being conducted in a single, trauma level I centre in Stockholm it
+is unrealistic to assume applicability of these findings at other
+hospitals facing unique challenges [@world2009guidelines]. It is previously known that
+low and middle income countries seem to benefit from certain AFs that
+high income countries do not and vice versa [@berg2022perceived].
 
@@
 
-With the ongoing academic, political, and public attention towards
-patient safety and quality of healthcare [@evans2009audit], it is
-important to discover optimal methods for monitoring and enhancing the
-care provided within trauma systems.
-
 Given the expanding field of artificial intelligence and machine
 learning, newer studies should focus on using these tools to screen for
-patient cases at risk of OFIs. This will diminish misclassification
-rates and ensure consistent and sustained high performance, attributable
+patient cases at risk of an OFI. This will diminish misclassification
+rates and ensure sustained high performance, attributable
 to an algorithm's capability to adapt and learn from the prevailing
 context of trauma care. This approach is contemporary but not
-unexplored. A preprint is currently available by Attergrim et al.
-[@attergrim2023predicting] exploring the idea of machine learning models
+unexplored. A recent preprint by Attergrim et al.
+[@attergrim2023predicting] explored the idea of machine learning models
 as a screening tool for potential OFIs. They found that machine learning
@@
 
-Conclusion
-==========
+# Conclusion
 
-AFs are widely used as a tool in quality improvement programmes and
+AFs are widely used as a tool in QI programmes and
 should reflect the current gap in trauma care quality. Strong evidence
@@
 
-References
-==========
+# References

### 2024-06-16 manuscript.Rmd
update

@@
 Trauma, defined as a physical injury and the body's associated response,
-accounts for 8-9% of global deaths annually [@world2014injuries;
-@roth2018global], mainly affecting younger individuals in low- and
-middle-income countries attributed to mainly road traffic injuries
-[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Furthermore,
-trauma is responsible for the top cause for disability-adjusted
-life-years (DALY) [@vos2020global] and a higher propensity towards
-mental disorders and cognitive decline [@kessler1995posttraumatic;
+causes 8-9% of global deaths annually [@world2014injuries;
+@roth2018global], disproportionately affecting younger individuals in low- and
+middle-income countries, primarily due to road traffic injuries
+[@vos2020global; @roth2018global; @SweTrau2022Arsrapport].
+Trauma also ranks highest in disability-adjusted life-years (DALYs) [@vos2020global] and increases the risk of mental disorders and cognitive decline [@kessler1995posttraumatic;
 @north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
-@visser2017course]. This negatively affects individuals, communities and
-society as a whole [@newnam2014impacts].It is estimated that the cost of
-road traffic injuries alone accounts for 2% of gross domestic product
+@visser2017course], impacting individuals, communities and
+society at large [@newnam2014impacts]. Road traffic injuries alone cost approximately 2% of gross domestic product
 (GDP) in high-income countries [@world2014injuries].
@@
 ## Setting
+Karolinska University Hospital, a Trauma level I
+hospital, treats around 1,500 trauma patients annually. If a patient case results in team activation or the
+patient has an Injury Severity Score (ISS) of less than 9
+retrospectively, it is added to the Karolinska
+trauma registry which reports to SweTrau. The registry includes data on vital
+signs, times, injuries, and interventions as well as patient
+demographics according to the European Consensus Statement, the Utstein
+template [@ringdal2008utstein]
 
+The Karolinska trauma registry also contains a local care quality
+database with specific AFs and OFIs that are determined in MMC
+conferences through consensus decisions. OFIs are assessed in multiple stages,
+characterised by escalating levels of scrutiny. Notably, instances of
+mortality are directly referred to the multidisciplinary conference,
+where, in addition to assessing OFIs, a decision is made regarding
+whether the death was preventable or potentially preventable, a
+classification also falling under the purview of OFIs.
+
+From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality.
+The review process was refined and formalized, with trauma cases initially assessed by a specialized trauma nurse for potential OFIs.
+In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses.
+Trauma quality data screened for AFs and cases flagged for possible care failures underwent secondary review by two specialized nurses.
+Potential OFIs identified in this review prompted a comprehensive evaluation of the trauma patient's case.
 ```{=html}
@@
 ## Participants
+
+The study included 8,309 patients treated at Karolinska University Hospital from 2012 to 2022 who underwent OFI screening.
+Patients under 15 and those not screened for OFIs were excluded due to differences in clinical management for minors.
 
@@
 ## Variables and data sources/measurements
+The outcome variable in this study is an OFI, as established by the MMCs
+and furthermore treated as a dichotomous variable with "Yes - At least
+one OFI identified" and "No - No OFI identified".
+The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1.
+Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.
 
+The data spanning 2012 to 2022 was retrieved from the
+Karolinska University Hospital trauma registry and trauma care quality
+database. Patient data on vital signs, care processes and interventions, level of
+care and time aspects was retrieved from SweTrau while both exposures
+(AFs) and outcome (OFI) were retrieved from the Karolinska University
+Hospital trauma care quality database. All data was anonymised to protect patient privacy.
 ```{=html}
@@
 ## Bias
-
+Consensus decisions at MMC conferences may introduce misclassification bias.
+Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias.
+Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias.
+Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.
 <!-- Describe any efforts to address potential sources of bias -->
@@
 ## Statistical methods
+The study utilized R [@R2022language] for statistical analysis.
+Patient characteristics between cases with and without OFI were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.
 
+The predictive performance of AFs was assessed by calculating sensitivity and specificity for each AF in predicting OFI.
+Receiver operating characteristic (ROC) curves were constructed for binary classification, and the area under the curve (AUC) was computed to compare AF performance.
+[@robin2011proc].
+
+Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and AUC.
+The significance of each AF's AUC was evaluated against chance (AUC = 0.5) and the minimum standard for diagnostic tests (AUC = 0.8) [@nahm2022receiver] using an empirical bootstrap test (1,000 iterations).
+
+Missing values for AFs were treated as negative (i.e., "not flagged").
 ```{=html}
@@
 Figure 1: Flowchart of the patient selection process. Shows the patients
-excluded and included in this study. <!--
+excluded and included in this study.
+
+<!--
 
@@
 ##Interpretation
-
-Higher ISS, longer times to CT, longer times to first intervention and
-higher levels of care were correlated with OFIs, corresponding with a
-recent study conducted at Karolinska University Hospital by Albaaj et
-al. [@albaaj2023patient]. Notably, GCS at ED showed a correlation with OFI even though the median
-GCS and its IQR was identical for individuals with OFI and without. Upon
-further analysis this could be explained by the difference in
-distribution outside the IQR.
+Higher ISS, longer times to CT, longer times to first intervention, and higher levels of care correlated with OFIs, consistent with a recent study by Albaaj et al.
+[@albaaj2023patient].
+Notably, GCS at the ED also showed a correlation with OFIs despite identical median GCS values, likely due to distribution differences outside the IQR.
 
 The AF "GCS\<9 and not intubated" showed no significant ability to
-predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]
-in 2008. An explanation for this could be that an abnormal GCS as a
-result from intoxication versus head injury for example are not
-comparable, suggesting that the AF might be more reflective of clinical
-decision rather than quality of care [@willis2008interpreting]. However,
-according to Sanddal et al. [@sanddal2011analysis] airway management was
-associated with OFI, suggesting that an alternative AF related to airway
-management might perform differently.
+predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]. Possibly because the AF might be more reflective of clinical
+decision rather than quality of care [@willis2008interpreting]. For instance, 
+An abnormal GCS resulting from intoxication is managed differently than one caused by a head injury.
 
-Both AFs measuring delay: "\>60 until first intervention" and "\>30min
-until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] found a correlation between delay
-and preventable death. Furthermore, Ghorbani et al.
-[@ghorbani2018analysis] identified delay in treatment and delay to CT as
-common errors in trauma care. A more recent study in Japan by Yamamoto
-et. al [@yamamoto2023immediate] showed a significant reduction in
-mortality when patients received a whole-body CT within 10 minutes of
-arrival at the ED. Notably, OFI in this study was not soley a reflection
-of mortality but also morbidity. Rather, it could highlight differences in quality gaps
-between hospitals.
+Both delay-related AFs: "\>60 until first intervention" and "\>30min
+until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] linked delays
+to preventable death and Ghorbani et al.
+[@ghorbani2018analysis] identified treatment and CT delays as common errors in trauma care. A recent study in Japan by Yamamoto
+et. al [@yamamoto2023immediate] found reduced mortality with whole-body CT scans within 10 minutes of ED arrival.
 
@@
 
-AFs, although widely used, there is lack
-of consensus on what they are, what purpose they serve and how they
-shall be used [@stelfox2010quality; @stelfox2011evidence;
-@berg2022perceived]. Further, AFs are static in their criteria, contrary to health care in general.
-Consequently, the utility of AFs may diminish with emerging challenges
-or upon resolution of the quality gaps they were designed for. This
-highlights the need for consistent and evidence-based review to
-determine current usefulness [@santana2014development]. However, such
-review process might present with costs greater than the potential
-benefit from AFs.
+AFs are widely used, but lack consensus on which to use, purpose, and application [@stelfox2010quality; @stelfox2011evidence;
+@berg2022perceived].
+They have static criteria and may not adapt to evolving healthcare challenges, or upon resolution of the quality gaps they were designed for, reducing their utility over time.
+This underscores the need for regular evidence-based reviews of AF performance [@santana2014development].
+However, these reviews might be cost-prohibitive
 
-While AFs may help determining OFIs, they could overshadow other factors
-contributing to that analysis, resulting in over-reliance. AFs might
-disproportionally affect certain groups of people failing to recognise
-patient cases that fall outside the scope of AF criteria
-[@willis2008interpreting]. For example, AFs primarily focus on aspects
-of trauma care mostly relevant to medical doctors despite the
-interdisciplinary nature of trauma care. This is especially
-counterintuitive given the broad definition of OFIs. Lastly, confounding
-variables adds to the uncertainty of AFs performance. For example,
-patients receiving the same interventions at different time points may
-do so as a function of injury severity not quality of care
+While AFs can help identify an OFI, they might overshadow other contributing factors, leading to over-reliance and potentially failing to recognize cases outside their scope.
+[@willis2008interpreting].
+For example, AFs often focus on aspects relevant to doctors, despite the interdisciplinary nature of trauma care.
+Lastly, Confounding variables, such as different treatment times based on injury severity, add to the uncertainty of AFs' effectiveness.
 [@willis2008interpreting].
@@
 is unrealistic to assume applicability of these findings at other
-hospitals facing unique challenges [@world2009guidelines]. It is previously known that
-low and middle income countries seem to benefit from certain AFs that
-high income countries do not and vice versa [@berg2022perceived].
+hospitals facing unique challenges [@world2009guidelines]. For instance AFs that benefit
+low and middle income may not benefit
+high income countries and vice versa [@berg2022perceived].
 
@@
 Given the expanding field of artificial intelligence and machine
-learning, newer studies should focus on using these tools to screen for
-patient cases at risk of an OFI. This will diminish misclassification
-rates and ensure sustained high performance, attributable
-to an algorithm's capability to adapt and learn from the prevailing
-context of trauma care. This approach is contemporary but not
-unexplored. A recent preprint by Attergrim et al.
-[@attergrim2023predicting] explored the idea of machine learning models
-as a screening tool for potential OFIs. They found that machine learning
+learning, new studies should focus on using these tools to screen for
+patient cases at risk of an OFI. This will reduce misclassification
+and maintain high performance, attributable
+to an algorithm's capability to adapt. A recent preprint by Attergrim et al.
+[@attergrim2023predicting] found that machine learning
 models outperformed the AFs currently used at Karolinska University
-Hospital in predicting OFIs.
-
-Should the use of AFs remain, it is important to ensure its performance
-by routinely updating them in order to better reflect the current
+Hospital in predicting OFIs. Should the use of AFs remain, they must be routinely updated to reflect current
 quality gaps. 
@@
 # References
+
+Keep? However,
+according to Sanddal et al. [@sanddal2011analysis] airway management was
+associated with OFI, suggesting that an alternative AF related to airway
+management might perform differently.
+
+Notably, OFI in this study was not soley a reflection
+of mortality but also morbidity. Rather, it could highlight differences in quality gaps
+between hospitals.

### 2024-06-17 manuscript.Rmd
update

@@
 ---
-title: "Performance of Individual Audit Filters in Predicting Opportunities for Improvement
-  in Adult Trauma Patients"
-subtitle: A single-center registry-based retrospective cohort
-output: html_document
+title: Performance of Individual Audit Filters in Predicting Opportunities for Improvement
+  in Adult Trauma Patients
 author: "Josef Al-Khalili"
+output:
+  html_document: default
+  pdf_document: default
+subtitle: "A single-center registry-based retrospective cohort"
 bibliography: bibliography.bib
@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
+source("main.R")
+library(rofi)
+library(dplyr)
+library(ggplot2)
+library(pROC)
+library(gtsummary)
+library(gt)
+library(boot)
+library(webshot2)
+library(knitr)
 ```
-
 # Abstract
@@
 [@world2009guidelines]. These programs form the core of current trauma
-care systems and their development around the world
-[@American2022resources]. QI programs consist of a series of recommended
-techniques aiming to facilitate discussion and analysis of the trauma
-process in order to take corrective action for future success. The
-following techniques are commonly used: a multidisciplinary mortality
-and morbidity conference (MMC), a preventable death review panel, and
+care systems
+[@American2022resources] and commonly include multidisciplinary mortality
+and morbidity conferences (MMC), preventable death review panels, and
 the use of AFs also known as quality indicators [@world2009guidelines].
-
-Trauma care is a multidisciplinary, complex and time-critical health
-care service provided by hospitals, often trauma centres.
-[@American2022resources]. Trauma centres have a significantly lower
-mortality rate compared to non-trauma centres, and the maturity of a
-trauma centre correlates with lower mortality, demonstrating the
-importance of high-quality, specialised trauma care
-[@alharbi2021effectiveness; @mann1999systematic;
-@candefjord2020mortality]. Despite this, even established trauma care
-systems are prone to errors due to the complex nature of trauma and its
-susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
-@o2013opportunities]. Studies between 1990 and 2014 have shown a pooled
-preventable death rate of 20%, with more recent studies showing a lower
-preventable death rate [@kwon2014pooled]. For example, a recent study in
+Despite QI programs, trauma care remains prone to errors [@nikouline2021errors; @vioque2014classifying;
+@o2013opportunities]. Studies have shown a pooled preventable death rate of 20% from 1990 to 2014, with more recent studies indicating lower rates [@kwon2014pooled]. For instance, a recent study in
 a Swedish Level I trauma centre found a preventable death rate of 4%
@@
 ### Morbidity and Mortality Conferences
-
-The MMC is a central QI technique. It is a proven framework with the
-critical components of anonymity, focusing on specific adverse events,
-recognising flawed approaches, critical analysis, improvement proposals
-and implementing such [@gregor2016morbidity].
-
-The conference participants are healthcare professionals involved in
-trauma care and its processes, but not necessarily the specific patient
-case. The selection of patient cases varies, but common themes for
-discussion are unexpected mortality, unexpected morbidity and errors
-[@orlander2003morbidity].
-
-Implemented effectively, MMCs serve as an important tool for error
-reduction without being perceived as negative feedback for individual
-care providers, but rather as an incentive for reporting. These
-conferences promote a more comprehensive, systems-based approach that
-emphasises task analysis, teamwork and quality improvement
-[@lazzara2022morbidity]. The introduction of MMC has been shown to
-increase morbidity and mortality reporting [@kong2016analysis].
+The MMC is a central QI technique focusing on anonymity, critical analysis of specific adverse events, recognizing flawed approaches, and proposing and implementing improvements [@gregor2016morbidity].
+Attended by healthcare professionals involved in trauma care, MMCs discuss themes like unexpected mortality, morbidity, and errors [@orlander2003morbidity], fostering error reduction and encouraging reporting without negative feedback [@lazzara2022morbidity].
+Effective MMC implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].
 
@@
 
-AFs, also known as quality indicators are established criteria based on
-standardised care. Any deviation from the AFs should raise the attention
+AFs are criteria based on standardized care to enhance trauma care quality. Deviation from AFs should raise the attention
 of health care professionals with the purpose of improving the quality
-of care [@American2022resources]. An AF can involve outcomes such as
-"death" or criteria related to guidelines and gold standards such as
-"the placement of two large bore intravenous lines within 15 minutes
-from arrival to a health care facility" [@world2009guidelines].
+of care [@American2022resources].
+AFs can involve outcomes like "death" or adherence to guidelines, such as "placement of two large bore intravenous lines within 15 minutes of arrival." [@world2009guidelines].Studies have shown mixed results on the effectiveness of AFs in reducing trauma-related mortality.
+While some early studies reported mortality reduction [@chadbunchachai2001comparative;
+@chadbunchachai2003study], a 2009 review by Evans et al.
+[@evans2009audit] found no substantial evidence supporting AFs' effectiveness.
+Stelfox et al.
+[@stelfox2010quality] identified over 1,500 AFs but noted a lack of a common, evidence-based set for evaluating trauma care quality, despite the high perceived usefulness of AFs [@berg2022perceived].
 
-Previous studies have shown a reduction in trauma-related mortality
-after the implementation of AFs [@chadbunchachai2001comparative;
-@chadbunchachai2003study]. Per contra, newer studies seem to differ. In
-2009, an extensive review by Evans et al. [@evans2009audit] sought to
-determine if AFs could be used in improving processes in trauma care.
-They found no studies successfully providing evidence for the use of
-AFs. More than 1,500 AFs have been identified in a review by Stelfox et
-al. [@stelfox2010quality]. Notably, they did not find a common set of
-clearly defined, evidence-based, broadly accepted AFs for evaluating the
-quality of trauma care. The perceived usefulness of audit filters still
-remains high [@berg2022perceived]
-
-Opportunities for improvement (OFIs) are the endpoint of patient care.
-Following the review of individual patient cases there is a decision
-made regarding the existence of an OFI [@American2022resources]. It has
-been shown that such review processes are associated with high-quality
-trauma care [@santana2014development]. OFIs are typically associated
-with failures in initial care [@vioque2014classifying], specifically in
+Opportunities for improvement (OFIs) are decided upon review of individual patient cases [@American2022resources]. They are typically associated
+with failures in initial care [@vioque2014classifying] like
 airway management, fluid resuscitation, haemorrhage control and chest
 injury management [@sanddal2011analysis; @roy2017learning;
-@o2013opportunities].
-
-There is lacking evidence for the use of AFs for improving trauma care
+@o2013opportunities]. There is lacking evidence for the use of AFs for improving trauma care
 [@evans2009audit; @stelfox2010quality] and the focus has been on
 mortality as outcome measure [@chadbunchachai2001comparative;
-@chadbunchachai2003study]. Furthermore, the use of AFs has also been
-associated with high frequencies of false positive OFI predictions,
+@chadbunchachai2003study]. Furthermore, the use of AFs has been
+associated with high frequencies of false positives,
 ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
 some AFs do not appear to correlate with OFIs at all [@lewis2017delay].
+
+The aim of this study is to determine the performance of AFs in
+predicting OFIs and furthermore assess their relevance as a screening
+tool for MMCs.
 
@@
 ## Variables and data sources/measurements
-The outcome variable in this study is an OFI, as established by the MMCs
-and furthermore treated as a dichotomous variable with "Yes - At least
+The outcome variable is an OFI, as established by the MMCs
+and treated as a dichotomous variable with "Yes - At least
 one OFI identified" and "No - No OFI identified".
@@
 Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.
-<!-- Describe any efforts to address potential sources of bias -->
 
@@
 between hospitals.
+Trauma care is a multidisciplinary, complex and time-critical health
+care service provided by hospitals, often trauma centres.
+[@American2022resources]. Trauma centres have a significantly lower
+mortality rate compared to non-trauma centres, and the maturity of a
+trauma centre correlates with lower mortality, demonstrating the
+importance of high-quality, specialised trauma care
+[@alharbi2021effectiveness; @mann1999systematic;
+@candefjord2020mortality]. 
+ It has
+been shown that such review processes are associated with high-quality
+trauma care [@santana2014development].

### 2024-06-18 manuscript.Rmd
update

@@
 ---
+
 ```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
@@
 library(boot)
-library(webshot2)
 library(knitr)
 ```
+
 # Abstract
@@
 
-Trauma is responsible for 9% of global deaths annually
-and has a significant impact on individuals and society. Quality
-Improvement (QI) programs seek to minimize these adverse outcomes using
-Audit Filters (AFs) and interdisciplinary Morbidity and Mortality
-Conferences (MMC). AFs are based on criteria for standard care and can
-function as a screening tool of eligible patient cases for discussion at
-MMCs. One of the goals with MMCs is to determine potential Opportunities
-for Improvement (OFI).This study aimed to evaluate the
-performance of individual AFs in predicting OFIs and to assess their
-relevance as a screening tool for Morbidity and Mortality Conferences
-(MMC).
+Trauma is responsible for 9% of global deaths annually and has a
+significant impact on individuals and society. Quality Improvement (QI)
+programs seek to minimize these adverse outcomes using Audit Filters
+(AFs) and interdisciplinary Morbidity and Mortality Conferences (MMC).
+AFs are based on criteria for standard care and can function as a
+screening tool of eligible patient cases for discussion at MMCs. One of
+the goals with MMCs is to determine potential Opportunities for
+Improvement (OFI).This study aimed to evaluate the performance of
+individual AFs in predicting OFIs and to assess their relevance as a
+screening tool for Morbidity and Mortality Conferences (MMC).
 
@@
 
-Register data from 8,309 patient cases
-were collected for analysis. The sensitivity, specificity and Area Under
-the Receiver Operating Characteristics Curve (AUC) was calculated to
-assess performance. An empirical bootstrapping test of AUC values was
-conducted to compare the performance of each AF to coincidence (AUC =
-0.5) and minimum AUC for diagnostic tests (AUC = 0.8).
+Register data from 8,309 patient cases were collected for analysis. The
+sensitivity, specificity and Area Under the Receiver Operating
+Characteristics Curve (AUC) was calculated to assess performance. An
+empirical bootstrapping test of AUC values was conducted to compare the
+performance of each AF to coincidence (AUC = 0.5) and minimum AUC for
+diagnostic tests (AUC = 0.8).
 
@@
 
-Four
-out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
-AFs demonstrated a correlation with OFI, with AUC ranging from 0.51 to
-0.63. The AF "\> 60 min until first intervention" exhibited the highest
-AUC (0.63, 95% CI [0.61-0.65]).
+Four out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The
+remaining AFs demonstrated a correlation with OFI, with AUC ranging from
+0.51 to 0.63. The AF "\> 60 min until first intervention" exhibited the
+highest AUC (0.63, 95% CI [0.61-0.65]).
 
@@
 
-No individual AF seemed to
-perform well in predicting OFI, which suggest rethinking the current
-screening method for MMCs.
+No individual AF seemed to perform well in predicting OFI, which suggest
+rethinking the current screening method for MMCs.
 
@@
 causes 8-9% of global deaths annually [@world2014injuries;
-@roth2018global], disproportionately affecting younger individuals in low- and
-middle-income countries, primarily due to road traffic injuries
-[@vos2020global; @roth2018global; @SweTrau2022Arsrapport].
-Trauma also ranks highest in disability-adjusted life-years (DALYs) [@vos2020global] and increases the risk of mental disorders and cognitive decline [@kessler1995posttraumatic;
-@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
-@visser2017course], impacting individuals, communities and
-society at large [@newnam2014impacts]. Road traffic injuries alone cost approximately 2% of gross domestic product
-(GDP) in high-income countries [@world2014injuries].
+@roth2018global], disproportionately affecting younger individuals in
+low- and middle-income countries, primarily due to road traffic injuries
+[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Trauma also
+ranks highest in disability-adjusted life-years (DALYs) [@vos2020global]
+and increases the risk of mental disorders and cognitive decline
+[@kessler1995posttraumatic; @north1999psychological;
+@muscatelli2017prevalence; @newnam2014impacts; @visser2017course],
+impacting individuals, communities and society at large
+[@newnam2014impacts]. Road traffic injuries alone cost approximately 2%
+of gross domestic product (GDP) in high-income countries
+[@world2014injuries].
 
@@
 [@world2009guidelines]. These programs form the core of current trauma
-care systems
-[@American2022resources] and commonly include multidisciplinary mortality
-and morbidity conferences (MMC), preventable death review panels, and
-the use of AFs also known as quality indicators [@world2009guidelines].
-Despite QI programs, trauma care remains prone to errors [@nikouline2021errors; @vioque2014classifying;
-@o2013opportunities]. Studies have shown a pooled preventable death rate of 20% from 1990 to 2014, with more recent studies indicating lower rates [@kwon2014pooled]. For instance, a recent study in
-a Swedish Level I trauma centre found a preventable death rate of 4%
+care systems [@American2022resources] and commonly include
+multidisciplinary mortality and morbidity conferences (MMC), preventable
+death review panels, and the use of AFs also known as quality indicators
+[@world2009guidelines]. Despite QI programs, trauma care remains prone
+to errors [@nikouline2021errors; @vioque2014classifying;
+@o2013opportunities]. Studies have shown a pooled preventable death rate
+of 20% from 1990 to 2014, with more recent studies indicating lower
+rates [@kwon2014pooled]. For instance, a recent study in a Swedish Level
+I trauma centre found a preventable death rate of 4%
 [@ghorbani2018analysis].
@@
 ### Morbidity and Mortality Conferences
-The MMC is a central QI technique focusing on anonymity, critical analysis of specific adverse events, recognizing flawed approaches, and proposing and implementing improvements [@gregor2016morbidity].
-Attended by healthcare professionals involved in trauma care, MMCs discuss themes like unexpected mortality, morbidity, and errors [@orlander2003morbidity], fostering error reduction and encouraging reporting without negative feedback [@lazzara2022morbidity].
-Effective MMC implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].
+
+The MMC is a central QI technique focusing on anonymity, critical
+analysis of specific adverse events, recognizing flawed approaches, and
+proposing and implementing improvements [@gregor2016morbidity]. Attended
+by healthcare professionals involved in trauma care, MMCs discuss themes
+like unexpected mortality, morbidity, and errors
+[@orlander2003morbidity], fostering error reduction and encouraging
+reporting without negative feedback [@lazzara2022morbidity]. Effective
+MMC implementation has been shown to increase morbidity and mortality
+reporting [@kong2016analysis].
 
@@
 
-AFs are criteria based on standardized care to enhance trauma care quality. Deviation from AFs should raise the attention
-of health care professionals with the purpose of improving the quality
-of care [@American2022resources].
-AFs can involve outcomes like "death" or adherence to guidelines, such as "placement of two large bore intravenous lines within 15 minutes of arrival." [@world2009guidelines].Studies have shown mixed results on the effectiveness of AFs in reducing trauma-related mortality.
-While some early studies reported mortality reduction [@chadbunchachai2001comparative;
-@chadbunchachai2003study], a 2009 review by Evans et al.
-[@evans2009audit] found no substantial evidence supporting AFs' effectiveness.
-Stelfox et al.
-[@stelfox2010quality] identified over 1,500 AFs but noted a lack of a common, evidence-based set for evaluating trauma care quality, despite the high perceived usefulness of AFs [@berg2022perceived].
+AFs are criteria based on standardized care to enhance trauma care
+quality. Deviation from AFs should raise the attention of health care
+professionals with the purpose of improving the quality of care
+[@American2022resources]. AFs can involve outcomes like "death" or
+adherence to guidelines, such as "placement of two large bore
+intravenous lines within 15 minutes of arrival."
+[@world2009guidelines].Studies have shown mixed results on the
+effectiveness of AFs in reducing trauma-related mortality. While some
+early studies reported mortality reduction
+[@chadbunchachai2001comparative; @chadbunchachai2003study], a 2009
+review by Evans et al. [@evans2009audit] found no substantial evidence
+supporting AFs' effectiveness. Stelfox et al. [@stelfox2010quality]
+identified over 1,500 AFs but noted a lack of a common, evidence-based
+set for evaluating trauma care quality, despite the high perceived
+usefulness of AFs [@berg2022perceived].
 
-Opportunities for improvement (OFIs) are decided upon review of individual patient cases [@American2022resources]. They are typically associated
-with failures in initial care [@vioque2014classifying] like
+Opportunities for improvement (OFIs) are decided upon review of
+individual patient cases [@American2022resources]. They are typically
+associated with failures in initial care [@vioque2014classifying] like
 airway management, fluid resuscitation, haemorrhage control and chest
 injury management [@sanddal2011analysis; @roy2017learning;
-@o2013opportunities]. There is lacking evidence for the use of AFs for improving trauma care
-[@evans2009audit; @stelfox2010quality] and the focus has been on
-mortality as outcome measure [@chadbunchachai2001comparative;
-@chadbunchachai2003study]. Furthermore, the use of AFs has been
-associated with high frequencies of false positives,
-ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
-some AFs do not appear to correlate with OFIs at all [@lewis2017delay].
+@o2013opportunities]. There is lacking evidence for the use of AFs for
+improving trauma care [@evans2009audit; @stelfox2010quality] and the
+focus has been on mortality as outcome measure
+[@chadbunchachai2001comparative; @chadbunchachai2003study]. Furthermore,
+the use of AFs has been associated with high frequencies of false
+positives, ranging from 24% to 80% [@sanddal2011analysis;
+@roy2017learning], and some AFs do not appear to correlate with OFIs at
+all [@lewis2017delay].
 
@@
 
+We conducted a registry-based study using all trauma patients included
+in both the Karolinska University Hospital trauma registry and trauma
+care quality database between 2012 and 2022 to compare the performance
+of AFs in their ability to predict OFI. This study was approved by
+Stockholm Research Ethics Review Board, approval number 2021-02541 and
+2021-03531.
+
 <!-- Present key elements of study design -->
@@
 ## Setting
-Karolinska University Hospital, a Trauma level I
-hospital, treats around 1,500 trauma patients annually. If a patient case results in team activation or the
-patient has an Injury Severity Score (ISS) of less than 9
-retrospectively, it is added to the Karolinska
-trauma registry which reports to SweTrau. The registry includes data on vital
-signs, times, injuries, and interventions as well as patient
-demographics according to the European Consensus Statement, the Utstein
-template [@ringdal2008utstein]
+
+Karolinska University Hospital, a Trauma level I hospital, treats around
+1,500 trauma patients annually. If a patient case results in team
+activation or the patient has an Injury Severity Score (ISS) of less
+than 9 retrospectively, it is recorded in the Karolinska trauma
+registry, which reports to SweTrau. This registry includes data on vital
+signs, timings, injuries, interventions, and patient demographics as per
+the European Consensus Statement (Utstein template)
+[@ringdal2008utstein]
 
 The Karolinska trauma registry also contains a local care quality
-database with specific AFs and OFIs that are determined in MMC
-conferences through consensus decisions. OFIs are assessed in multiple stages,
-characterised by escalating levels of scrutiny. Notably, instances of
-mortality are directly referred to the multidisciplinary conference,
-where, in addition to assessing OFIs, a decision is made regarding
-whether the death was preventable or potentially preventable, a
+database with AFs and OFIs, identified by MMCs through consensus
+decisions. OFIs are assessed in multiple stages of scrutiny. Notably,
+mortality is directly referred to the MMCs in order to assess OFIs and
+deciding whether the death was preventable or potentially preventable, a
 classification also falling under the purview of OFIs.
 
-From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality.
-The review process was refined and formalized, with trauma cases initially assessed by a specialized trauma nurse for potential OFIs.
-In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses.
-Trauma quality data screened for AFs and cases flagged for possible care failures underwent secondary review by two specialized nurses.
-Potential OFIs identified in this review prompted a comprehensive evaluation of the trauma patient's case.
+From 2013 to 2017, efforts focused on
+identifying adverse outcomes unrelated to mortality. Initially, trauma cases where individually assessed by a
+specialized trauma nurse for potential OFIs. In 2017, this procedure
+became standard practice, incorporating preliminary evaluations by
+specialized trauma nurses. Cases
+flagged by an AF or identified by the nurse as potential
+care failures underwent a second review by two specialized nurses. Potential OFIs identified in this review
+prompted review in a MMC.
+
 ```{=html}
@@
 
-The study included 8,309 patients treated at Karolinska University Hospital from 2012 to 2022 who underwent OFI screening.
-Patients under 15 and those not screened for OFIs were excluded due to differences in clinical management for minors.
+The study included 8,309 patients treated at Karolinska University
+Hospital from 2012 to 2022 who underwent OFI screening. Patients under
+15 and those not screened for OFIs were excluded due to differences in
+clinical management for minors.
 
@@
 ## Variables and data sources/measurements
-The outcome variable is an OFI, as established by the MMCs
-and treated as a dichotomous variable with "Yes - At least
-one OFI identified" and "No - No OFI identified".
-The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1.
-Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.
 
-The data spanning 2012 to 2022 was retrieved from the
-Karolinska University Hospital trauma registry and trauma care quality
-database. Patient data on vital signs, care processes and interventions, level of
+The outcome variable is an OFI, as established by the MMCs and treated
+as a dichotomous variable with "Yes - At least one OFI identified" and
+"No - No OFI identified". The 11 AFs used at Karolinska University
+Hospital served as the exposure variable, as shown in Table 1. Each AF
+is labeled "original" or "manually created." Manually created AFs were
+calculated using trauma registry data, while original AFs were recorded
+in the quality database based on criteria beyond the available registry
+data.
+
+The data spanning 2012 to 2022 was retrieved from the Karolinska
+University Hospital trauma registry and trauma care quality database.
+Patient data on vital signs, care processes and interventions, level of
 care and time aspects was retrieved from SweTrau while both exposures
 (AFs) and outcome (OFI) were retrieved from the Karolinska University
-Hospital trauma care quality database. All data was anonymised to protect patient privacy.
+Hospital trauma care quality database. All data was anonymised to
+protect patient privacy.
+
 ```{=html}
@@
 ## Bias
-Consensus decisions at MMC conferences may introduce misclassification bias.
-Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias.
-Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias.
-Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.
 
-## Study size
-
-<!-- Explain how the study size was arrived at -->
-
-## Quantitative variables
+Consensus decisions at MMC conferences may introduce misclassification
+bias. Additionally, AF flagging by specialized nurses is a manual
+process, potentially leading to selection bias. Healthcare professionals
+attending MMCs may not accurately recall key details impacting OFI
+decisions, introducing recall bias. Furthermore, professionals not
+present for specific cases might over- or underestimate OFI
+significance.
 
@@
 ## Statistical methods
-The study utilized R [@R2022language] for statistical analysis.
-Patient characteristics between cases with and without OFI were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.
 
-The predictive performance of AFs was assessed by calculating sensitivity and specificity for each AF in predicting OFI.
-Receiver operating characteristic (ROC) curves were constructed for binary classification, and the area under the curve (AUC) was computed to compare AF performance.
-[@robin2011proc].
+The study utilized R [@R2022language] for statistical analysis. Patient
+characteristics between cases with and without OFI were compared using
+the Wilcoxon signed-rank test for continuous variables and the
+Chi-squared test for categorical variables.
 
-Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and AUC.
-The significance of each AF's AUC was evaluated against chance (AUC = 0.5) and the minimum standard for diagnostic tests (AUC = 0.8) [@nahm2022receiver] using an empirical bootstrap test (1,000 iterations).
+The predictive performance of AFs was assessed by calculating
+sensitivity and specificity for each AF in predicting OFI. Receiver
+operating characteristic (ROC) curves were constructed for binary
+classification, and the area under the curve (AUC) was computed to
+compare AF performance. [@robin2011proc].
+
+Bootstrap resampling (1,000 iterations) was applied to estimate 95%
+confidence intervals for sensitivity, specificity, and AUC. The
+significance of each AFs' AUC was evaluated against chance (AUC = 0.5)
+and the minimum standard for diagnostic tests (AUC = 0.8)
+[@nahm2022receiver] using an empirical bootstrap test (1,000
+iterations).
 
 Missing values for AFs were treated as negative (i.e., "not flagged").
+
 ```{=html}
@@
 
-##Participants After excluding patients under 15 years old and patients
-not screened for OFIs, a total of 8,309 individuals were included in the
+##Participants 
+After excluding patients under 15 years old and patients not screened for OFIs, a total of 8,309 individuals were included in the
 study. 7,797 out of 8,309 (93.8%) patient cases did not have an OFI and
@@
 
+```{=html}
 <!--
@@
 -->
-
+```
 You can include code in this document like this:
@@
 
-The aim of this study was to evaluate the performance of individual AFs in predicting OFIs and 
-to assess their suitability as a screening tool for MMCs.
-This study showed that no AFs was significantly successful in predicting OFIs.
-Four AFs ("SBP<90," "dead at 30 days," "GCS<9 and not intubated," and "CPR and thoracotomy") were equivalent to random chance in their predictive ability.
-The highest AUC was for ">60 minutes until first intervention,".
-Notably, all AFs performed significantly below the accepted standard for diagnostic tests (>0.8).
+The aim of this study was to evaluate the performance of individual AFs
+in predicting OFIs and to assess their suitability as a screening tool
+for MMCs. This study showed that no AFs was significantly successful in
+predicting OFIs. Four AFs ("SBP\<90," "dead at 30 days," "GCS\<9 and not
+intubated," and "CPR and thoracotomy") were equivalent to random chance
+in their predictive ability. The highest AUC was for "\>60 minutes until
+first intervention,". Notably, all AFs performed significantly below the
+accepted standard for diagnostic tests (\>0.8).
 
@@
 
-##Interpretation
-Higher ISS, longer times to CT, longer times to first intervention, and higher levels of care correlated with OFIs, consistent with a recent study by Albaaj et al.
-[@albaaj2023patient].
-Notably, GCS at the ED also showed a correlation with OFIs despite identical median GCS values, likely due to distribution differences outside the IQR.
+##Interpretation Higher ISS, longer times to CT, longer times to first
+intervention, and higher levels of care correlated with OFIs, consistent
+with a recent study by Albaaj et al. [@albaaj2023patient]. Notably, GCS
+at the ED also showed a correlation with OFIs despite identical median
+GCS values, likely due to distribution differences outside the IQR.
 
 The AF "GCS\<9 and not intubated" showed no significant ability to
-predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]. Possibly because the AF might be more reflective of clinical
-decision rather than quality of care [@willis2008interpreting]. For instance, 
-An abnormal GCS resulting from intoxication is managed differently than one caused by a head injury.
+predict OFIs, similarly shown by Willis et al.
+[@willis2008interpreting]. Possibly because the AF might be more
+reflective of clinical decision rather than quality of care
+[@willis2008interpreting]. For instance, An abnormal GCS resulting from
+intoxication is managed differently than one caused by a head injury.
 
 Both delay-related AFs: "\>60 until first intervention" and "\>30min
-until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] linked delays
-to preventable death and Ghorbani et al.
-[@ghorbani2018analysis] identified treatment and CT delays as common errors in trauma care. A recent study in Japan by Yamamoto
-et. al [@yamamoto2023immediate] found reduced mortality with whole-body CT scans within 10 minutes of ED arrival.
+until first CT", showed a correlation with OFI. Comparably, Teixeira et
+al. [@teixeira2007preventable] linked delays to preventable death and
+Ghorbani et al. [@ghorbani2018analysis] identified treatment and CT
+delays as common errors in trauma care. A recent study in Japan by
+Yamamoto et. al [@yamamoto2023immediate] found reduced mortality with
+whole-body CT scans within 10 minutes of ED arrival.
 
-"Dead at 30 days"
-and "CPR and thoracotomy" showed no correlation with OFI. This could be
-explained by inevitable mortality, thus presenting with a smaller time
-frame for mistakes to be made.
+"Dead at 30 days" and "CPR and thoracotomy" showed no correlation with
+OFI. This could be explained by inevitable mortality, thus presenting
+with a smaller time frame for mistakes to be made.
 
-AFs
-generally exhibited a high specificity relative to their sensitivity
+AFs generally exhibited a high specificity relative to their sensitivity
 which may be attributed to the outcome being rather uncommon, whilst
@@
 the AFs are too rigid in their criteria. On the contrary, a too
-inclusive criteria might flag several patients, without an OFI [@willis2008interpreting].
+inclusive criteria might flag several patients, without an OFI
+[@willis2008interpreting].
 
@@
 
-AFs are widely used, but lack consensus on which to use, purpose, and application [@stelfox2010quality; @stelfox2011evidence;
-@berg2022perceived].
-They have static criteria and may not adapt to evolving healthcare challenges, or upon resolution of the quality gaps they were designed for, reducing their utility over time.
-This underscores the need for regular evidence-based reviews of AF performance [@santana2014development].
-However, these reviews might be cost-prohibitive
+AFs are widely used, but lack consensus on which to use, purpose, and
+application [@stelfox2010quality; @stelfox2011evidence;
+@berg2022perceived]. They have static criteria and may not adapt to
+evolving healthcare challenges, or upon resolution of the quality gaps
+they were designed for, reducing their utility over time. This
+underscores the need for regular evidence-based reviews of AF
+performance [@santana2014development]. However, these reviews might be
+cost-prohibitive
 
-While AFs can help identify an OFI, they might overshadow other contributing factors, leading to over-reliance and potentially failing to recognize cases outside their scope.
-[@willis2008interpreting].
-For example, AFs often focus on aspects relevant to doctors, despite the interdisciplinary nature of trauma care.
-Lastly, Confounding variables, such as different treatment times based on injury severity, add to the uncertainty of AFs' effectiveness.
-[@willis2008interpreting].
+While AFs can help identify an OFI, they might overshadow other
+contributing factors, leading to over-reliance and potentially failing
+to recognize cases outside their scope. [@willis2008interpreting]. For
+example, AFs often focus on aspects relevant to doctors, despite the
+interdisciplinary nature of trauma care. Lastly, Confounding variables,
+such as different treatment times based on injury severity, add to the
+uncertainty of AFs' effectiveness. [@willis2008interpreting].
 
-Due to the
-study being conducted in a single, trauma level I centre in Stockholm it
-is unrealistic to assume applicability of these findings at other
-hospitals facing unique challenges [@world2009guidelines]. For instance AFs that benefit
-low and middle income may not benefit
-high income countries and vice versa [@berg2022perceived].
+Due to the study being conducted in a single, trauma level I centre in
+Stockholm it is unrealistic to assume applicability of these findings at
+other hospitals facing unique challenges [@world2009guidelines]. For
+instance AFs that benefit low and middle income may not benefit high
+income countries and vice versa [@berg2022perceived].
 
@@
 learning, new studies should focus on using these tools to screen for
-patient cases at risk of an OFI. This will reduce misclassification
-and maintain high performance, attributable
-to an algorithm's capability to adapt. A recent preprint by Attergrim et al.
-[@attergrim2023predicting] found that machine learning
-models outperformed the AFs currently used at Karolinska University
-Hospital in predicting OFIs. Should the use of AFs remain, they must be routinely updated to reflect current
-quality gaps. 
-<!--
+patient cases at risk of an OFI. This will reduce misclassification and
+maintain high performance, attributable to an algorithm's capability to
+adapt. A recent preprint by Attergrim et al. [@attergrim2023predicting]
+found that machine learning models outperformed the AFs currently used
+at Karolinska University Hospital in predicting OFIs. Should the use of
+AFs remain, they must be routinely updated to reflect current quality
+gaps. <!--
 
@@
 
-AFs are widely used as a tool in QI programmes and
-should reflect the current gap in trauma care quality. Strong evidence
-supporting their efficacy in quality improvement is scarce. In this
-study, no audit filter seemed to show promising performance in
-predicting OFIs with AUC ranging from 0.5 to 0.63. These results
-highlight the importance of continually evaluating the relevance of AFs.
-However, this process takes time and resources and using an alternative
-to AFs is recommended.
+AFs are widely used as a tool in QI programmes and should reflect the
+current gap in trauma care quality. Strong evidence supporting their
+efficacy in quality improvement is scarce. In this study, no audit
+filter seemed to show promising performance in predicting OFIs with AUC
+ranging from 0.5 to 0.63. These results highlight the importance of
+continually evaluating the relevance of AFs. However, this process takes
+time and resources and using an alternative to AFs is recommended.
 
@@
 
-Keep? However,
-according to Sanddal et al. [@sanddal2011analysis] airway management was
-associated with OFI, suggesting that an alternative AF related to airway
-management might perform differently.
+Keep? However, according to Sanddal et al. [@sanddal2011analysis] airway
+management was associated with OFI, suggesting that an alternative AF
+related to airway management might perform differently.
 
-Notably, OFI in this study was not soley a reflection
-of mortality but also morbidity. Rather, it could highlight differences in quality gaps
-between hospitals.
-Trauma care is a multidisciplinary, complex and time-critical health
-care service provided by hospitals, often trauma centres.
-[@American2022resources]. Trauma centres have a significantly lower
-mortality rate compared to non-trauma centres, and the maturity of a
-trauma centre correlates with lower mortality, demonstrating the
+Notably, OFI in this study was not soley a reflection of mortality but
+also morbidity. Rather, it could highlight differences in quality gaps
+between hospitals. Trauma care is a multidisciplinary, complex and
+time-critical health care service provided by hospitals, often trauma
+centres. [@American2022resources]. Trauma centres have a significantly
+lower mortality rate compared to non-trauma centres, and the maturity of
+a trauma centre correlates with lower mortality, demonstrating the
 importance of high-quality, specialised trauma care
 [@alharbi2021effectiveness; @mann1999systematic;
-@candefjord2020mortality]. 
- It has
-been shown that such review processes are associated with high-quality
-trauma care [@santana2014development].
+@candefjord2020mortality]. It has been shown that such review processes
+are associated with high-quality trauma care [@santana2014development].

### 2024-06-18 functions/confidence_interval.R

@@ -1,6 +1,6 @@
 confidence_interval <- function(twoVariableData){
   set.seed(123)
-  numberOfBootstraps <- 1000
+  numberOfBootstraps <- 100
   bootResults <- boot(twoVariableData,statistic = boot_strap, R = numberOfBootstraps)
   boot2 <- boot(twoVariableData,statistic = boot_strap_AUC, R = numberOfBootstraps)
   AUCCI <- boot.ci(boot2, type = "basic")

### 2024-06-21 manuscript.Rmd
update

@@
 output:
+  word_document: default
+  pdf_document: default
   html_document: default
-  pdf_document: default
 subtitle: "A single-center registry-based retrospective cohort"
@@
 
-### Morbidity and Mortality Conferences
-
 The MMC is a central QI technique focusing on anonymity, critical
@@
 reporting [@kong2016analysis].
-
-### Audit Filters and Opportunities for Improvement
 
@@
 vs
-`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`,
+`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)`,
 `r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.

### 2024-06-21 functions/confidence_interval.R

@@ -1,6 +1,6 @@
 confidence_interval <- function(twoVariableData){
   set.seed(123)
-  numberOfBootstraps <- 100
+  numberOfBootstraps <- 1000
   bootResults <- boot(twoVariableData,statistic = boot_strap, R = numberOfBootstraps)
   boot2 <- boot(twoVariableData,statistic = boot_strap_AUC, R = numberOfBootstraps)
   AUCCI <- boot.ci(boot2, type = "basic")

### 2024-07-13 manuscript.Rmd
update

@@
 
-Trauma is responsible for 9% of global deaths annually and has a
-significant impact on individuals and society. Quality Improvement (QI)
-programs seek to minimize these adverse outcomes using Audit Filters
-(AFs) and interdisciplinary Morbidity and Mortality Conferences (MMC).
-AFs are based on criteria for standard care and can function as a
-screening tool of eligible patient cases for discussion at MMCs. One of
-the goals with MMCs is to determine potential Opportunities for
-Improvement (OFI).This study aimed to evaluate the performance of
-individual AFs in predicting OFIs and to assess their relevance as a
-screening tool for Morbidity and Mortality Conferences (MMC).
+Trauma audit filters, based on standards of care, serve as screening tools for patient cases to be discussed at interdisciplinary mortality and morbidity conferences.
+These conferences aim to identify potential opportunities for improvement in the care of trauma patients.
+This study aimed to assess the effectiveness of individual audit filters in predicting opportunities for improvement and their relevance as screening tools for morbidity and mortality conferences.
 
@@
 
-Register data from 8,309 patient cases were collected for analysis. The
-sensitivity, specificity and Area Under the Receiver Operating
-Characteristics Curve (AUC) was calculated to assess performance. An
-empirical bootstrapping test of AUC values was conducted to compare the
-performance of each AF to coincidence (AUC = 0.5) and minimum AUC for
-diagnostic tests (AUC = 0.8).
+We conducted a single-centre registry-based retrospective cohort study based on register data from 8,309 patient cases at the Karolinska University Hospital in Sweden.
+The sensitivity, specificity and Cohen’s kappa coefficient was calculated to assess performance.
+Bootstrapping of 1,000 iterations was used to calculate 95% confidence intervals.
 
@@
 
-Four out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The
-remaining AFs demonstrated a correlation with OFI, with AUC ranging from
-0.51 to 0.63. The AF "\> 60 min until first intervention" exhibited the
-highest AUC (0.63, 95% CI [0.61-0.65]).
+Our analysis included 8,309 patients.
+Two out of the 11 audit filters had a Cohen’s kappa coefficient of 0 and one audit filter had an inverse relation to opportunities for improvement.
+The remaining audit filters demonstrated a correlation with opportunities for improvement, with Cohen’s kappa coefficients ranging from 0.02 to 0.12.
+The audit filters “> 60 min until first intervention” and “ISS > 15 and not in ICU” exhibited the highest Cohen’s kappa coefficient (0.12, 95% CI [0.10-0.15]).
 
@@
 
-No individual AF seemed to perform well in predicting OFI, which suggest
-rethinking the current screening method for MMCs.
+No individual audit filter performed well in predicting opportunities for improvement, which suggest reassessing the current method of screening for mortality and morbidity conferences.
 
@@
 
-Trauma, defined as a physical injury and the body's associated response,
-causes 8-9% of global deaths annually [@world2014injuries;
-@roth2018global], disproportionately affecting younger individuals in
-low- and middle-income countries, primarily due to road traffic injuries
-[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Trauma also
-ranks highest in disability-adjusted life-years (DALYs) [@vos2020global]
-and increases the risk of mental disorders and cognitive decline
-[@kessler1995posttraumatic; @north1999psychological;
-@muscatelli2017prevalence; @newnam2014impacts; @visser2017course],
+Trauma, defined as a physical injury and the body’s associated response, causes 8-9% of global deaths annually [@world2014injuries;
+@roth2018global]and ranks highest in disability-adjusted life-years (DALYs) [@vos2020global],
 impacting individuals, communities and society at large
@@
 
-To reduce trauma burden, quality improvement (QI) programs have been
-established by the World Health Organization (WHO) and the International
-Association for Trauma Surgery and Intensive Care (IATSIC)
-[@world2009guidelines]. These programs form the core of current trauma
-care systems [@American2022resources] and commonly include
-multidisciplinary mortality and morbidity conferences (MMC), preventable
-death review panels, and the use of AFs also known as quality indicators
-[@world2009guidelines]. Despite QI programs, trauma care remains prone
-to errors [@nikouline2021errors; @vioque2014classifying;
-@o2013opportunities]. Studies have shown a pooled preventable death rate
-of 20% from 1990 to 2014, with more recent studies indicating lower
-rates [@kwon2014pooled]. For instance, a recent study in a Swedish Level
-I trauma centre found a preventable death rate of 4%
+Quality improvement programs form the core of current trauma care systems 
+[@American2022resources] and commonly include
+ techniques such as multidisciplinary mortality and morbidity conferences and preventable death review panels to improve trauma care quality
+[@world2009guidelines]. Despite these programs, trauma care remains prone to errors [@nikouline2021errors; @vioque2014classifying;
+@o2013opportunities].
+Studies have shown a pooled preventable death rate of 20%, with more recent studies indicating lower rates [@kwon2014pooled].
+A recent study in a Swedish Level I trauma centre found a preventable death rate of 4%
 [@ghorbani2018analysis].
 
-The MMC is a central QI technique focusing on anonymity, critical
-analysis of specific adverse events, recognizing flawed approaches, and
-proposing and implementing improvements [@gregor2016morbidity]. Attended
-by healthcare professionals involved in trauma care, MMCs discuss themes
-like unexpected mortality, morbidity, and errors
+The mortality and morbidity conferences is a central technique focusing on anonymity, critical analysis of specific adverse events and recognizing flawed approaches followed by proposing and implementing improvements [@gregor2016morbidity].
+Attended by healthcare professionals involved in trauma care, mortality and morbidity conferences discuss themes such as unexpected mortality, morbidity and errors
 [@orlander2003morbidity], fostering error reduction and encouraging
-reporting without negative feedback [@lazzara2022morbidity]. Effective
-MMC implementation has been shown to increase morbidity and mortality
-reporting [@kong2016analysis].
+reporting without negative feedback [@lazzara2022morbidity].
+Effective mortality and morbidity conferences implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].
 
-AFs are criteria based on standardized care to enhance trauma care
-quality. Deviation from AFs should raise the attention of health care
-professionals with the purpose of improving the quality of care
-[@American2022resources]. AFs can involve outcomes like "death" or
-adherence to guidelines, such as "placement of two large bore
-intravenous lines within 15 minutes of arrival."
-[@world2009guidelines].Studies have shown mixed results on the
-effectiveness of AFs in reducing trauma-related mortality. While some
-early studies reported mortality reduction
+Audit filters are used to identify cases for mortality and morbidity conferences.
+These filters are criteria based on standards of care, guidelines or outcomes and function as a screening tool for potential opportunities for improvement.
+Deviation from audit filters should raise the attention of health care providers with the purpose of improving the quality of care
+[@American2022resources]. Audit filters can involve outcomes such as “death” or adherence to guidelines, such as “placement of two large bore intravenous lines within 15 minutes of arrival.”
+[@world2009guidelines].Studies have shown conflicting results on the effectiveness of audit filters in reducing trauma-related mortality.
+While some early studies reported a mortality reduction following the implementation of audit filters
 [@chadbunchachai2001comparative; @chadbunchachai2003study], a 2009
-review by Evans et al. [@evans2009audit] found no substantial evidence
-supporting AFs' effectiveness. Stelfox et al. [@stelfox2010quality]
-identified over 1,500 AFs but noted a lack of a common, evidence-based
-set for evaluating trauma care quality, despite the high perceived
-usefulness of AFs [@berg2022perceived].
+review by Evans et al. [@evans2009audit]  found no high quality evidence supporting the effectiveness of audit filters. Furthermore, Stelfox et al.  [@stelfox2010quality]
+identified over 1,500 audit filters and noted a lack of a common, evidence-based set for evaluating trauma care quality.
+Despite the uncertainty, the perceived usefulness of audit filters remains high [@berg2022perceived].
 
-Opportunities for improvement (OFIs) are decided upon review of
-individual patient cases [@American2022resources]. They are typically
-associated with failures in initial care [@vioque2014classifying] like
-airway management, fluid resuscitation, haemorrhage control and chest
-injury management [@sanddal2011analysis; @roy2017learning;
-@o2013opportunities]. There is lacking evidence for the use of AFs for
-improving trauma care [@evans2009audit; @stelfox2010quality] and the
-focus has been on mortality as outcome measure
-[@chadbunchachai2001comparative; @chadbunchachai2003study]. Furthermore,
-the use of AFs has been associated with high frequencies of false
-positives, ranging from 24% to 80% [@sanddal2011analysis;
-@roy2017learning], and some AFs do not appear to correlate with OFIs at
-all [@lewis2017delay].
+Opportunities for improvement are decided upon review of individual patient cases and aim to reflect current trauma care quality gaps  [@American2022resources].
+These opportunities are typically associated with failures in initial care [@vioque2014classifying]  like airway management, fluid resuscitation, haemorrhage control and chest injury management [@sanddal2011analysis; @roy2017learning;
+@o2013opportunities]. There is lacking evidence for the use of audit filters for improving trauma care [@evans2009audit; @stelfox2010quality], but the focus has been on mortality as outcome measure 
+[@chadbunchachai2001comparative; @chadbunchachai2003study].Moreover, the use of audit filters  has been associated with high frequencies of false positives, ranging from 24% to 80% [@sanddal2011analysis;
+@roy2017learning], and some audit filters do not appear to correlate with opportunities for improvement at all [@lewis2017delay].
 
-The aim of this study is to determine the performance of AFs in
-predicting OFIs and furthermore assess their relevance as a screening
-tool for MMCs.
+The aim of this study is to determine the performance of audit filters in predicting opportunities for improvement and assessing their relevance as a screening tool for mortality and morbidity conferences
 
@@
 
-We conducted a registry-based study using all trauma patients included
-in both the Karolinska University Hospital trauma registry and trauma
-care quality database between 2012 and 2022 to compare the performance
-of AFs in their ability to predict OFI. This study was approved by
-Stockholm Research Ethics Review Board, approval number 2021-02541 and
-2021-03531.
-
-<!-- Present key elements of study design -->
+We conducted a single-centre registry-based retrospective cohort study  of all trauma patients included in both the Karolinska University Hospital trauma registry and trauma care quality database between 2012 and 2022 to compare the performance of audit filters in their ability to predict opportunities for improvement.
+This study was approved by Swedish Ethical Review Authority, approval number 2021-02541 and 2021-03531.
 
@@
 
-Karolinska University Hospital, a Trauma level I hospital, treats around
-1,500 trauma patients annually. If a patient case results in team
-activation or the patient has an Injury Severity Score (ISS) of less
-than 9 retrospectively, it is recorded in the Karolinska trauma
-registry, which reports to SweTrau. This registry includes data on vital
-signs, timings, injuries, interventions, and patient demographics as per
-the European Consensus Statement (Utstein template)
+Karolinska University Hospital manages around 1,500 trauma patients annually.
+The Karolinska Trauma Registry, which reports to the national trauma registry SweTrau, includes all patient cases that result in trauma team activation or who have an Injury Severity Score (ISS) of more than 9.
+This registry includes data on vital signs, timings, injuries, interventions, and patient demographics as per a European consensus statement
 [@ringdal2008utstein]
 
-The Karolinska trauma registry also contains a local care quality
-database with AFs and OFIs, identified by MMCs through consensus
-decisions. OFIs are assessed in multiple stages of scrutiny. Notably,
-mortality is directly referred to the MMCs in order to assess OFIs and
-deciding whether the death was preventable or potentially preventable, a
-classification also falling under the purview of OFIs.
+The trauma centre at the Karolinska University Hospital also keeps a local care quality database with information intended for the mortality and morbidity conferences, including identified triggered audit filters and opportunities for improvement.
+Opportunities for improvement are assessed in multiple stages of scrutiny.
+Notably, mortality is directly referred to the mortality and morbidity conferences to assess opportunities for improvement.
 
-From 2013 to 2017, efforts focused on
-identifying adverse outcomes unrelated to mortality. Initially, trauma cases where individually assessed by a
-specialized trauma nurse for potential OFIs. In 2017, this procedure
-became standard practice, incorporating preliminary evaluations by
-specialized trauma nurses. Cases
-flagged by an AF or identified by the nurse as potential
-care failures underwent a second review by two specialized nurses. Potential OFIs identified in this review
-prompted review in a MMC.
+From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality.
+Initially, trauma cases were individually assessed by a specialized trauma nurse for potential opportunities for improvement.
+In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses.
+Cases flagged by an audit filter or identified by the nurse as potential care failures underwent a second review by two specialized nurses.
+Potential opportunities for improvement identified in this review prompted review by the mortality and morbidity conference.
 
@@
 
-The study included 8,309 patients treated at Karolinska University
-Hospital from 2012 to 2022 who underwent OFI screening. Patients under
-15 and those not screened for OFIs were excluded due to differences in
-clinical management for minors.
+The study included all patients treated at Karolinska University Hospital from 2012 to 2022 who underwent screening for opportunities for improvement.
+Patients under 15 were excluded due to differences in clinical management.
 
@@
 ## Variables and data sources/measurements
+The outcome was opportunities for improvement, as established by the mortality and morbidity conference, and was treated dichotomously as “Yes - At least one opportunity for improvement identified” and “No - No opportunity for improvement identified”.
 
-The outcome variable is an OFI, as established by the MMCs and treated
-as a dichotomous variable with "Yes - At least one OFI identified" and
-"No - No OFI identified". The 11 AFs used at Karolinska University
-Hospital served as the exposure variable, as shown in Table 1. Each AF
-is labeled "original" or "manually created." Manually created AFs were
-calculated using trauma registry data, while original AFs were recorded
-in the quality database based on criteria beyond the available registry
-data.
+The eleven audit filters used at the Karolinska University Hospital served as the exposure variables, as shown in Table 1.
+The data spanning 2012 to 2022 was retrieved from the Karolinska Trauma Registry and trauma care quality database.
+Patient data on vital signs, care processes and interventions, level of care and time aspects was retrieved from SweTrau while both the exposure audit filters and the outcome opportunities for improvement were retrieved from the quality database.
 
-The data spanning 2012 to 2022 was retrieved from the Karolinska
-University Hospital trauma registry and trauma care quality database.
-Patient data on vital signs, care processes and interventions, level of
-care and time aspects was retrieved from SweTrau while both exposures
-(AFs) and outcome (OFI) were retrieved from the Karolinska University
-Hospital trauma care quality database. All data was anonymised to
-protect patient privacy.
+All data was anonymised to protect patient privacy.
 
@@
 
-Consensus decisions at MMC conferences may introduce misclassification
-bias. Additionally, AF flagging by specialized nurses is a manual
-process, potentially leading to selection bias. Healthcare professionals
-attending MMCs may not accurately recall key details impacting OFI
-decisions, introducing recall bias. Furthermore, professionals not
-present for specific cases might over- or underestimate OFI
-significance.
-
+Consensus decisions at the mortality and morbidity conferences may introduce misclassification bias.
+Additionally, audit filters flagging by specialized nurses is a manual process, potentially leading to selection bias.
+Healthcare professionals attending mortality and morbidity conferences may not accurately recall key details impacting opportunities for improvement decisions, introducing recall bias.
+Furthermore, professionals, not present for specific cases might over- or underestimate opportunities for improvement significance.
 ```{=html}
@@
 
-The study utilized R [@R2022language] for statistical analysis. Patient
-characteristics between cases with and without OFI were compared using
-the Wilcoxon signed-rank test for continuous variables and the
-Chi-squared test for categorical variables.
+ for statistical analysis.
+Patient characteristics for cases with and without opportunities for improvement were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.
 
-The predictive performance of AFs was assessed by calculating
-sensitivity and specificity for each AF in predicting OFI. Receiver
-operating characteristic (ROC) curves were constructed for binary
-classification, and the area under the curve (AUC) was computed to
-compare AF performance. [@robin2011proc].
+The predictive performance of audit filters was assessed by calculating the sensitivity, specificity and Cohen’s kappa for each audit filter in predicting opportunities for improvement.
+Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and Cohen’s kappa coefficient.
 
-Bootstrap resampling (1,000 iterations) was applied to estimate 95%
-confidence intervals for sensitivity, specificity, and AUC. The
-significance of each AFs' AUC was evaluated against chance (AUC = 0.5)
-and the minimum standard for diagnostic tests (AUC = 0.8)
-[@nahm2022receiver] using an empirical bootstrap test (1,000
-iterations).
+Missing values for audit filters were treated as negative (i.e., “not flagged”).
 
-Missing values for AFs were treated as negative (i.e., "not flagged").
-
-```{=html}
-<!--
-
-(a) Describe all statistical methods, including those used to control
-for confounding
-
-(b) Describe any methods used to examine subgroups and interactions
-
-(c) Explain how missing data were addressed
-
-(d)
-
-Cohort study: If applicable, explain how loss to follow-up was addressed
-
-Case-control study:If applicable, explain how matching of cases and
-controls was addressed
-
-Cross-sectional study: If applicable, describe analytical methods
-taking account of sampling strategy
-
-(e) Describe any sensitivity analyses
--->
-```
 # Results
@@
 ##Participants 
-After excluding patients under 15 years old and patients not screened for OFIs, a total of 8,309 individuals were included in the
-study. 7,797 out of 8,309 (93.8%) patient cases did not have an OFI and
-512 out of 8,309 (6.2%) had an OFI. Figure 1 presents a flowchart
-showcasing the exclusion and inclusion of the patient cases from the
-trauma registry.
+After excluding patients under 15 years old and patients not screened for opportunities for improvement, a total of 8,309 individuals were included in the study.
+7,797 out of 8,309 (93.8%) patient cases did not have an opportunity for improvement and 512 out of 8,309 (6.2%) had an opportunity for improvement, see Figure 1.
 
-Most trauma patients were men (n = 5,755, 69%). Patient characteristics
-varied depending on whether patients had an OFI or not. Respectively,
-the mean age was significantly higher\
-`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
-`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
-`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")` and
-the ISS was significantly higher
-`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
-`r inline_text(tableOne, variable = ISS, column = stat_1)`,
-`r inline_text(tableOne, variable = ISS, column = "p.value")`. The
-number of intubations in the emergency department was significantly
-higher
-`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
-vs
-`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)`,
+Patients with an opportunity for improvement had significantly higher mean age `r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs `r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
+`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`, ISS `r inline_text(tableOne, variable = ISS, column = stat_2)` vs
+`r inline_text(tableOne, variable = ISS, column = stat_1)`, `r inline_text(tableOne, variable = ISS, column = "p.value")` and intubations in the emergency department (ED) `r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
+vs `r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)`,
 `r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
-Patients with OFI had longer times to definitive treatment from hospital
-arrival compared to patients without OFI
-`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
-vs
+Patients with opportunity for improvement had longer times to definitive treatment from hospital arrival `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)` vs
 `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`,
 `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
-The time to the first CT was also longer in patients with OFIs compared
-to patients without
+ and longer times to first CT
 `r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
 `r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`,
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`.
-
-The number of OFIs were highest in patients admitted to the intensive
-care unit
-`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)`
-and surgical ward
-`r inline_text(tableOne, variable = host_care_level, level = "Surgical ward", column = stat_2)`.
-
-Gender `r inline_text(tableOne, variable = Gender, column = "p.value")`,
-the mean systolic blood pressure
-`r inline_text(tableOne, variable = ed_sbp_value, column = "p.value")`,
-death at 30 days
-`r inline_text(tableOne, variable = res_survival, column = "p.value")`
-and trauma team activation
-`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was
-not significantly correlated with OFI.
-
-Notably, there was a statistical significant correlation between GCS at
-ED and OFI (p \< 0.001).
-
-Table 2 shows the missing data for each AF. As mentioned in the method
-section, these values were interpreted as "not flagged" with an AF. The
-variables with the highest amount of missing data were "time to
-definitive treatment" (n = 5,990, 72.1%) and "time to first CT" (n =
-1,012, 12.2%). The smallest amount of missing data was found in
-"SBP\<90" (n = 13, 0.2%).
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`. The number of opportunities for improvement were highest in patients admitted to the intensive care unit 
+`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)` compared to general ward `r inline_text(tableOne, variable = host_care_level, level = "General ward", column = stat_2)`.
+Gender, mean systolic blood pressure, death at 30 days and trauma team activation was not significantly correlated with opportunities for improvement (p>0.05).
+Notably, there was a statistically significant correlation between GCS at ED and opportunities for improvement (p < 0.001).
+The highest amount of missing data was observed for the audit filter "time to
+definitive treatment" (n = 5,990, 72.1%). The smallest amount of missing data was found in "SBP\<90" (n = 13, 0.2%).
 
@@
 
-The performance of each AF was determined through a calculation of
-sensitivity and specificity which is demonstrated in Table 4.
+The number of patient cases flagged by each individual audit filter varied between 107 for “GCS<9 and not intubated” and 3907 for “>30 min until CT”.
+The highest number of missing values was observed with “>60 min until first intervention” (n = 5990, 72.1%).
+The lowest number of missing values was observed with “dead at 30 days” (n = 11, 0.1%) and “ISS > 15 and not in ICU” (n = 11, 0.1%).
+See table 2 for missing values for each audit filter.
 
-The number of patient cases flagged by each individual AF varied between
-107 for "GCS\<9 and not intubated" and 3907 for "\>30 min until CT". The
-AF with the highest sensitivity was "\>30 min until CT" (59.2%, 95% CI
-[55.2-63.8]). The lowest sensitivity was seen in "GCS\<9 and not
-intubated" (1.2%, 95% CI [0.2-2]). The AF with the highest specificity
-was "GCS\<9 and not intubated" (98.7%, 95% CI [98.5-98.9]). The lowest
-specificity was seen in "\>30 min until CT" (53.8%, 95% CI [52.7-54.8]).
-Notably, the confidence intervals varied between sensitivity and
-specificity, with sensitivity displaying a greater interval.
+The audit filter with the highest sensitivity was “>30 min until CT” (59.2%, 95% CI [55.2-63.8]).
+The lowest sensitivity was seen in “GCS<9 and not intubated” (1.2%, 95% CI [0.2-2]).
+The audit filter with the highest specificity was “GCS<9 and not intubated” (98.7%, 95% CI [98.5-98.9]).
+The lowest specificity was seen in “>30 min until CT” (53.8%, 95% CI [52.7-54.8]).
+Notably, the confidence intervals varied between sensitivity and specificity, with sensitivity displaying  wider intervals.
+See Table 4 for the performance of all audit filters using sensitivity and specificity.
 
-Table 5 summarises the performance of each individual AF in identifying
-OFIs in trauma care by AUC. Four of the 11 AFs, "SBP\<90", "dead at 30
-days", "GCS\<9 and not intubated" and "CPR and thoracotomy" showed a
-performance in predicting OFIs equal to that of coincidence (AUC = 0.5).
-The highest AUC was seen in "\>60 min until first intervention" (0.63,
-95% CI [0.61-0.65]). All AFs showed significantly lower AUC compared to
-the minimum standard for diagnostic tests (AUC = 0.8) (p \< 0.001).
-
-Table 2: Comparison of missing values for each AF. Missing values is
-shown as a number and percentage of the total number of patient cases
-included in the study (n = 8309).\
-`r tableTwo` Table 3: Patient characteristics of individuals with and
-without an OFI. Missing values are shown only when they exist.
-`r tableOne` Table 4: Comparison of performance for each AF in
-predicting OFIs. Performance is defined by sensitivity (%) and
-specificity (%). A 95% confidence interval is shown together with each
-performance indicator. The number of cases flagged by each filter is
-displayed.\
-`r tableThree` Table 5: Comparison of performance for each AF in
-predicting OFIs. Performance is defined by the area under the receiver
-operating characteristics curve. P-values show the significance of AUC
-compared to coincidence (AUC = 0.5) and to the minimum standard for
-diagnostic tests (AUC = 0.8) [@nahm2022receiver]. The p-values were
-calculated using an empirical bootstrap test. Furthermore, a 95%
-confidence interval was used. `r tableFour`
+Two of the eleven audit filters, “GCS < 9 and not intubated” and “CPR and thoracotomy”, had a Cohen’s kappa coefficient of 0.
+The highest Cohen’s kappa was seen for the audit filter “>60 min until first intervention” and “ISS > 15 and not in ICU” (0.12, 95% CI [0.10-0.15]).
+All audit filters showed none to slight correlation with opportunities for improvement.
+See table 5 for the performance of all audit filters using Cohen’s kappa coefficient.
 
@@
 ```
-
-Figure 1: Flowchart of the patient selection process. Shows the patients
-excluded and included in this study.
 
@@
 
-The aim of this study was to evaluate the performance of individual AFs
-in predicting OFIs and to assess their suitability as a screening tool
-for MMCs. This study showed that no AFs was significantly successful in
-predicting OFIs. Four AFs ("SBP\<90," "dead at 30 days," "GCS\<9 and not
-intubated," and "CPR and thoracotomy") were equivalent to random chance
-in their predictive ability. The highest AUC was for "\>60 minutes until
-first intervention,". Notably, all AFs performed significantly below the
-accepted standard for diagnostic tests (\>0.8).
+The aim of this study was to evaluate the performance of individual audit filters in predicting opportunities for improvement and to assess their suitability as a screening tool for mortality and morbidity conferences.
+This study showed that audit filters had “no” to “slight” ability to predict opportunities for improvement.
+Two audit filters, “GCS<9 and not intubated” and “CPR and thoracotomy”, showed no predictive ability.
+The highest Cohen’s kappa coefficient was seen for the audit filter “>60 min until first intervention” and “ISS > 15 and not in ICU”, but it was still low.
 
@@
 
-OFI, while dichotomously defined, includes a diverse set of outcomes.
-This convolutes the process of creating AFs that perform well in
-predicting OFIs. AFs would favour the identification of some, but not
-all, errors.
+Opportunity for improvement, while binarily defined, includes a diverse set of outcomes.
+This complicates the process of creating audit filters that perform well in predicting opportunities for improvement.
+Audit filters would favour the identification of some, but not all, errors.
 
-Each AF was individually related to OFI. Consequently, if an AF
-demonstrated a true positive correlation with an OFI, instances where
-the same patient case had multiple AFs introduce uncertainty regarding
-which specific AF truthfully predicted the OFI.
+Each audit filter was individually related to opportunities for improvement.
+Consequently, if an audit filter demonstrated a true positive correlation with an opportunity for improvement, instances where the same patient case had multiple audit filters introduce uncertainty regarding which specific audit filter truthfully predicted the opportunity for improvement.
 
-The number of missing values varied greatly among the AFs. As stated
-above, missing values from each AF were coded as "not flagged" by an AF.
-This means that some instances where patient cases were not flagged,
-might actually meet the AF criteria, but for some reason not documented.
-This reduces the validity of some AFs that featured high amounts of
-missing data.
+The number of missing values varied greatly among the audit filters.
+As stated above, missing values from each audit filter were coded as “not flagged” by an audit filter.
+This means that instances where patient cases were not flagged might meet the audit filter criteria, but were not documented.
+This reduces the validity of audit filters that feature high amounts of missing data.
 
-##Interpretation Higher ISS, longer times to CT, longer times to first
-intervention, and higher levels of care correlated with OFIs, consistent
-with a recent study by Albaaj et al. [@albaaj2023patient]. Notably, GCS
-at the ED also showed a correlation with OFIs despite identical median
-GCS values, likely due to distribution differences outside the IQR.
+##Interpretation
+Higher ISS, longer times to CT, longer times to first intervention, and higher levels of care correlated with opportunities for improvement, consistent with a recent Swedish study by Albaaj et al.[@albaaj2023patient].
+Notably, GCS at the ED also showed a correlation with opportunities for improvement despite identical median GCS values, likely due to distribution differences outside the interquartile range (IQR).
 
-The AF "GCS\<9 and not intubated" showed no significant ability to
-predict OFIs, similarly shown by Willis et al.
-[@willis2008interpreting]. Possibly because the AF might be more
-reflective of clinical decision rather than quality of care
-[@willis2008interpreting]. For instance, An abnormal GCS resulting from
-intoxication is managed differently than one caused by a head injury.
+The audit filter “GCS<9 and not intubated” showed no significant ability to predict opportunities for improvement, similarly shown by Willis et al. 
+[@willis2008interpreting]. Possibly because the audit filter might be more reflective of clinical decisions rather than quality of care 
+[@willis2008interpreting]. For instance, an abnormal GCS because of intoxication is managed differently to an abnormal GCS due to head injury.
 
-Both delay-related AFs: "\>60 until first intervention" and "\>30min
-until first CT", showed a correlation with OFI. Comparably, Teixeira et
-al. [@teixeira2007preventable] linked delays to preventable death and
-Ghorbani et al. [@ghorbani2018analysis] identified treatment and CT
-delays as common errors in trauma care. A recent study in Japan by
-Yamamoto et. al [@yamamoto2023immediate] found reduced mortality with
-whole-body CT scans within 10 minutes of ED arrival.
+Both delay-related audit filters: “>60 min until first intervention” and “>30min until first CT”, showed a slight correlation with opportunity for improvement.
+Comparably, Teixeira et al.[@teixeira2007preventable] linked delays to preventable death andidentified delay to treatment and delays to CT as common errors in trauma care.
+A recent study in Japan by Yamamoto et.
+al [@yamamoto2023immediate] observed significantly reduced mortality with whole-body CT scans within 10 minutes of arrival to the ED.
 
-"Dead at 30 days" and "CPR and thoracotomy" showed no correlation with
-OFI. This could be explained by inevitable mortality, thus presenting
-with a smaller time frame for mistakes to be made.
+“CPR and thoracotomy” showed no correlation with opportunities for improvement.
+This could be explained by non-preventable mortality, thus presenting a smaller time frame for mistakes to be made.
+A similar explanation could be given to the inverse relation between “Dead at 30 days” and opportunities for improvement.
 
-AFs generally exhibited a high specificity relative to their sensitivity
-which may be attributed to the outcome being rather uncommon, whilst
-every individual AF misses the majority of OFIs. This could suggest that
-the AFs are too rigid in their criteria. On the contrary, a too
-inclusive criteria might flag several patients, without an OFI
-[@willis2008interpreting].
+Audit filters generally exhibited a high specificity relative to their sensitivity which may be attributed to the outcome being rather uncommon, whilst every individual audit filter misses the majority of opportunities for improvement.
+This could suggest that the audit filters are too rigid in their criteria.
+On the contrary, too inclusive criteria might flag several patients, without opportunities for improvement [@willis2008interpreting].
 
@@
 
-AFs are widely used, but lack consensus on which to use, purpose, and
-application [@stelfox2010quality; @stelfox2011evidence;
-@berg2022perceived]. They have static criteria and may not adapt to
-evolving healthcare challenges, or upon resolution of the quality gaps
-they were designed for, reducing their utility over time. This
-underscores the need for regular evidence-based reviews of AF
-performance [@santana2014development]. However, these reviews might be
-cost-prohibitive
+Audit filters are widely used, but lack consensus on which to use, purpose, and application [@stelfox2010quality; @stelfox2011evidence;
+@berg2022perceived].
+They have static criteria and may therefore not adapt to evolving healthcare challenges, or upon resolution of the quality gaps they were designed for, reducing their utility over time.
+This underscores the need for regular evidence-based reviews of audit filter performance [@santana2014development].
+However, these reviews might be cost-prohibitive.
 
-While AFs can help identify an OFI, they might overshadow other
-contributing factors, leading to over-reliance and potentially failing
-to recognize cases outside their scope. [@willis2008interpreting]. For
-example, AFs often focus on aspects relevant to doctors, despite the
-interdisciplinary nature of trauma care. Lastly, Confounding variables,
-such as different treatment times based on injury severity, add to the
-uncertainty of AFs' effectiveness. [@willis2008interpreting].
+While audit filters can help identify an opportunity for improvement, they might overshadow other contributing factors, leading to over-reliance and potentially failing to recognize cases outside their scope.
+[@willis2008interpreting].For example, audit filters often focus on aspects relevant to doctors, despite the interdisciplinary nature of trauma care.
+Lastly, confounding variables, such as different treatment times based on injury severity, add to the uncertainty of audit filters’ effectiveness.
+[@willis2008interpreting].
 
-Due to the study being conducted in a single, trauma level I centre in
-Stockholm it is unrealistic to assume applicability of these findings at
-other hospitals facing unique challenges [@world2009guidelines]. For
-instance AFs that benefit low and middle income may not benefit high
-income countries and vice versa [@berg2022perceived].
+Due to the study being conducted in a single trauma centre in Stockholm it is unrealistic to assume applicability of these findings at other hospitals facing unique challenges [@world2009guidelines].
+For instance, audit filters that benefit low and middle income may not benefit high income countries and vice versa [@berg2022perceived].
 
@@
 
-Given the expanding field of artificial intelligence and machine
-learning, new studies should focus on using these tools to screen for
-patient cases at risk of an OFI. This will reduce misclassification and
-maintain high performance, attributable to an algorithm's capability to
-adapt. A recent preprint by Attergrim et al. [@attergrim2023predicting]
-found that machine learning models outperformed the AFs currently used
-at Karolinska University Hospital in predicting OFIs. Should the use of
-AFs remain, they must be routinely updated to reflect current quality
-gaps. <!--
-
-Key results
------------
-Summarise key results with reference to study objectives
-
-Limitations
------------
-
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias
-
-Interpretation
---------------
-
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from
-similar studies, and other relevant evidence
-
-Generalisability
-----------------
-
-Discuss the generalisability (external validity) of the study results
-
--->
-
 # Conclusion
 
-AFs are widely used as a tool in QI programmes and should reflect the
-current gap in trauma care quality. Strong evidence supporting their
-efficacy in quality improvement is scarce. In this study, no audit
-filter seemed to show promising performance in predicting OFIs with AUC
-ranging from 0.5 to 0.63. These results highlight the importance of
-continually evaluating the relevance of AFs. However, this process takes
-time and resources and using an alternative to AFs is recommended.
+Audit filters are widely used as a tool in quality improvement programmes and should reflect the current gap in trauma care quality.
+In this study, no audit filter showed a significant performance in predicting opportunities for improvement, with Cohen’s kappa coefficient ranging from -0.01 to 0.12.
+These results highlight the importance of continually evaluating the relevance of audit filters.
+
+# Acknowledgements
+The authors thank all professionals taking part in the monthly mortality and morbidity conferences.
 
 # References
-
-Keep? However, according to Sanddal et al. [@sanddal2011analysis] airway
-management was associated with OFI, suggesting that an alternative AF
-related to airway management might perform differently.
-
-Notably, OFI in this study was not soley a reflection of mortality but
-also morbidity. Rather, it could highlight differences in quality gaps
-between hospitals. Trauma care is a multidisciplinary, complex and
-time-critical health care service provided by hospitals, often trauma
-centres. [@American2022resources]. Trauma centres have a significantly
-lower mortality rate compared to non-trauma centres, and the maturity of
-a trauma centre correlates with lower mortality, demonstrating the
-importance of high-quality, specialised trauma care
-[@alharbi2021effectiveness; @mann1999systematic;
-@candefjord2020mortality]. It has been shown that such review processes
-are associated with high-quality trauma care [@santana2014development].

### 2024-07-13 functions/confidence_interval_sens_spec.R

@@ -1,6 +1,6 @@
 confidence_interval_sens_spec <- function(twoVariableData){
   set.seed(123) #Setting the seed for bootstrap
-  numberOfBootstraps <- 1000 #Determining number of bootstrap iterations
+  numberOfBootstraps <- 5 #Determining number of bootstrap iterations
   bootResults <- boot(twoVariableData,statistic = bootstrap_sens_spec, R = numberOfBootstraps) #bootstrapping 
   sensitivityCI <- boot.ci(bootResults, type = "basic", index = 1)  # Sensitivity CI
   specificityCI <- boot.ci(bootResults, type = "basic", index = 2) #calculating confidence interval from bootstrap

### 2024-07-13 functions/create_audit_filters.R

@@ -1,8 +1,16 @@
 create_audit_filters <- function(cleanData){
-  cleanData$AF_sap_less90 <- cleanData$ed_sbp_value < 90 | cleanData$ed_sbp_rtscat <= 3
+
+  cleanData$AF_sap_less90 <- ifelse(
+    is.na(cleanData$ed_sbp_value) & is.na(cleanData$ed_sbp_rtscat), NA, 
+    ifelse(is.na(cleanData$ed_sbp_value), cleanData$ed_sbp_rtscat <= 3, 
+           ifelse(cleanData$ed_sbp_value < 90 , TRUE,FALSE)))
+  #cleanData$AF_sap_less90 <- cleanData$ed_sbp_value < 90 | cleanData$ed_sbp_rtscat <= 3
   #cleanData$AF_sap_less90[is.na(cleanData$AF_sap_less90)] <- FALSE
   #VK_sap_less90
-  cleanData$AF_gcs_less9_ej_intubTE <- cleanData$ed_gcs_sum < 9 & cleanData$ed_intubated != "Yes"
+  cleanData$AF_gcs_less9_ej_intubTE <- ifelse(
+    is.na(cleanData$ed_gcs_sum) | is.na(cleanData$ed_intubated) | cleanData$ed_gcs_sum >= 9, NA, 
+    ifelse(cleanData$ed_gcs_sum < 9 & cleanData$ed_intubated != "Yes", TRUE,FALSE))
+  #cleanData$AF_gcs_less9_ej_intubTE <- cleanData$ed_gcs_sum < 9 & cleanData$ed_intubated != "Yes"
   #cleanData$AF_gcs_less9_ej_intubTE[is.na(cleanData$AF_gcs_less9_ej_intubTE)] <- FALSE
   #VK_gcs_less9_ej_intubTE
   cleanData$AF_mer_30min_DT <- cleanData$dt_ed_first_ct > 30
@@ -11,13 +19,19 @@ create_audit_filters <- function(cleanData){
   cleanData$AF_mer_60_min_interv <- cleanData$dt_ed_emerg_proc > 60
   #cleanData$AF_mer_60_min_interv[is.na(cleanData$AF_mer_60_min_interv)] <- FALSE
   #VK_mer_60_min_interv
-  cleanData$AF_iss_15_ej_iva <- cleanData$ISS >= 15 & cleanData$host_care_level != "Intensive care unit"
[…]

### 2024-07-13 main.R

@@ -53,6 +53,7 @@ importDataOfi <- import_data_ofi(data) #
 cleanData <- clean_data(importDataOfi)
 dataWithAuditFilters <- create_audit_filters(cleanData)
 selectedData <- select_data(dataWithAuditFilters)
+
 createFlowchart <- create_flowchart(importDataOfi)
 
 tableOneData <- select(selectedData, ofi, 
@@ -71,18 +72,20 @@ for(auditFilter in selectedAuditFilter){
   twoVariableData <- select(selectedData, auditFilter, ofi)
   totalValues <- nrow(twoVariableData)
   missingValue <- sum(is.na(twoVariableData[1]))
-  
-  twoVariableData[1][is.na(twoVariableData[1])] <- FALSE
-  numberOfTrue <- sum(twoVariableData[[1]])
-  print(sum(is.na(twoVariableData[1])))
-  print(missingValue)
+  #print(missingValue)
+  #print(twoVariableData)
+  twoVariableDataNaOmit <- na.omit(twoVariableData)
+  #twoVariableData[1][is.na(twoVariableData[1])] <- FALSE #turning Na values false.
+  numberOfTrue <- sum(twoVariableDataNaOmit[[1]])
+  #print(sum(is.na(twoVariableDataNaOmit[1])))
+  #print(missingValue)
   if (counter %in% c(1,2,3,5,6,7,8)){
     manualOrOriginal <- "manual"
   } else {
     manualOrOriginal <- "original"
   }
-  confidenceIntervalKappa <- confidence_interval_kappa(twoVariableData)
-  confidenceIntervalSensitivitySpecificity <- confidence_interval_sens_spec(twoVariableData)
+  confidenceIntervalKappa <- confidence_interval_kappa(twoVariableDataNaOmit)
+  confidenceIntervalSensitivitySpecificity <- confidence_interval_sens_spec(twoVariableDataNaOmit)
   paste(c(round(calculateAUC, digits = 2)," (",confidenceInterval[5],"-",confidenceInterval[6],") "), collapse = "")
   print(confidenceIntervalKappa)
   
@@ -141,7 +144,8 @@ tableOne <- tableOneData %>%
   add_p(test = list(all_continuous() ~ "wilcox.test")) %>%
   bold_p() 
 ############TABLE ONE#######################
-'''
+
+'
 tableThree <- tableOfCalculatedData1 %>%
   gt() %>% 
[…]

### 2024-07-15 functions/confidence_interval_sens_spec.R
update

@@ -1,6 +1,6 @@
 confidence_interval_sens_spec <- function(twoVariableData){
   set.seed(123) #Setting the seed for bootstrap
-  numberOfBootstraps <- 5 #Determining number of bootstrap iterations
+  numberOfBootstraps <- 1000 #Determining number of bootstrap iterations
   bootResults <- boot(twoVariableData,statistic = bootstrap_sens_spec, R = numberOfBootstraps) #bootstrapping 
   sensitivityCI <- boot.ci(bootResults, type = "basic", index = 1)  # Sensitivity CI
   specificityCI <- boot.ci(bootResults, type = "basic", index = 2) #calculating confidence interval from bootstrap

### 2024-07-15 functions/create_audit_filters.R

@@ -14,6 +14,7 @@ create_audit_filters <- function(cleanData){
   #cleanData$AF_gcs_less9_ej_intubTE[is.na(cleanData$AF_gcs_less9_ej_intubTE)] <- FALSE
   #VK_gcs_less9_ej_intubTE
   cleanData$AF_mer_30min_DT <- cleanData$dt_ed_first_ct > 30
+
   #cleanData$AF_mer_30min_DT[is.na(cleanData$AF_mer_30min_DT)] <- FALSE
   #VK_mer_30min_DT
   cleanData$AF_mer_60_min_interv <- cleanData$dt_ed_emerg_proc > 60
@@ -34,23 +35,42 @@ create_audit_filters <- function(cleanData){
     ifelse(cleanData$ISS >= 15 & cleanData$Tr_Nivå == "No", TRUE,FALSE))
   #cleanData$AF_iss_15_ej_TE[is.na(cleanData$AF_iss_15_ej_TE)] <- FALSE
   #ISS 15 och ingen trauma larm
-  cleanData$AF_ej_trombrof_TBI_72h <- cleanData$VK_ej_trombrof_TBI_72h == "ja"
+  
+  check_starts_with_1 <- function(row) {
+    any(grepl("^1", row))
+  }
+  AISCols <- grep("^AISCode_", names(cleanData))
+  cleanData$TBI <- apply(cleanData[AISCols], 1, function(row){
+    if (check_starts_with_1(row)) {
+      return(TRUE)
+    } else {
+      return(NA)
+    }
+  })
+  print(cleanData$TBI)
+  cleanData$AF_ej_trombrof_TBI_72h <- ifelse(is.na(cleanData$TBI) | is.na(cleanData$VK_ej_trombrof_TBI_72h), NA, 
+                                             ifelse(cleanData$TBI == TRUE & cleanData$VK_ej_trombrof_TBI_72h == "ja", TRUE, FALSE))
+  print(cleanData$AF_ej_trombrof_TBI_72h)
   #turning values into boolean
-  cleanData$AF_hlr_thorak <- cleanData$VK_hlr_thorak == "ja"
+  cleanData$AF_hlr_thorak <- ifelse(is.na(cleanData$VK_hlr_thorak) & (is.na(cleanData$Fr1.12) | cleanData$Fr1.12 != 1), NA, 
+                                    ifelse(cleanData$VK_hlr_thorak == "nej" | cleanData$Fr1.12 == 1, FALSE, TRUE))  
+
   #turning values into boolean
[…]

### 2024-07-15 main.R

@@ -18,16 +18,16 @@ listOfAuditFilters <- c("AF_sap_less90","AF_death_30d","AF_iss_15_ej_TE",
                         "AF_mass_transf", "AF_gcs_less9_ej_intubTE", 
                         "AF_iss_15_ej_iva", "AF_mer_60_min_interv",
                         "AF_mer_30min_DT", "AF_hlr_thorak", 
-                        "AF_lever_och_mjaltskada", "AF_ej_trombrof_TBI_72h")
+                        "AF_lever_och_mjaltskada", "AF_ej_trombrof_TBI_72h", "AF_all")
 listOfAuditFiltersClean <- c("SBP < 90", "Dead at 30 days", "ISS > 15 and no team activation", 
                              "Massive transfusion", "GCS < 9 and not intubated", "ISS > 15 and not in ICU", 
                              "> 60 min until first intervention", "> 30 min until first CT",
                              "CPR and thoracotomy", "Liver or spleen injury", 
-                             "No anticoagulants within 72 hours after TBI")
+                             "No anticoagulants within 72 hours after TBI", "All")
 
 #1,2,3,5,6,7,8 manual
 ## viktigt att det ska vara samma ordning ^
-selectedAuditFilter <- listOfAuditFilters[1:11]
+selectedAuditFilter <- listOfAuditFilters[1:12]
 
 tableKappa <- data.frame(Auditfilter = character(0),
                          Sensitivity = numeric(0),

### 2024-09-14 manuscript.Rmd
update

@@
 output:
+  html_document: default
   word_document: default
   pdf_document: default
-  html_document: default
 subtitle: "A single-center registry-based retrospective cohort"
@@
 library(knitr)
+library(webshot2)
 ```
@@
 
-Trauma audit filters, based on standards of care, serve as screening tools for patient cases to be discussed at interdisciplinary mortality and morbidity conferences.
-These conferences aim to identify potential opportunities for improvement in the care of trauma patients.
-This study aimed to assess the effectiveness of individual audit filters in predicting opportunities for improvement and their relevance as screening tools for morbidity and mortality conferences.
+Trauma audit filters, based on standards of care, serve as screening
+tools for patient cases to be assessed at interdisciplinary mortality
+and morbidity conferences. These conferences aim to identify potential
+opportunities for improvement in the care of trauma patients. This study
+aimed to determine the effectiveness of individual audit filters in
+predicting opportunities for improvement and their relevance as
+screening tools for morbidity and mortality conferences.
 
@@
 
-We conducted a single-centre registry-based retrospective cohort study based on register data from 8,309 patient cases at the Karolinska University Hospital in Sweden.
-The sensitivity, specificity and Cohen’s kappa coefficient was calculated to assess performance.
-Bootstrapping of 1,000 iterations was used to calculate 95% confidence intervals.
+We conducted a single-centre registry-based retrospective cohort study based on register data collected between 2013 and 2022 from the Karolinska University Hospital in Sweden.
+Performance was measured using sensitivity, specificity and Cohen’s kappa coefficient for ten audit filters.
+Bootstrapping with 1,000 iterations calculated 95% confidence intervals.
 
@@
 Our analysis included 8,309 patients.
-Two out of the 11 audit filters had a Cohen’s kappa coefficient of 0 and one audit filter had an inverse relation to opportunities for improvement.
-The remaining audit filters demonstrated a correlation with opportunities for improvement, with Cohen’s kappa coefficients ranging from 0.02 to 0.12.
-The audit filters “> 60 min until first intervention” and “ISS > 15 and not in ICU” exhibited the highest Cohen’s kappa coefficient (0.12, 95% CI [0.10-0.15]).
+Audit filters showed a Cohen’s kappa coefficient ranging from -0.07 (95% CI [-0.14-0]) for “GCS < 9 and not intubated” and 0.08 (95% CI [0.05-0.12]) for “Massive transfusion”.
+The Cohen’s kappa coefficient for all audit filters combined was 0.04 (95% CI[0.03-0.04]).
 
@@
 
-No individual audit filter performed well in predicting opportunities for improvement, which suggest reassessing the current method of screening for mortality and morbidity conferences.
+No individual audit filter performed well in predicting opportunities for improvement, which suggests reassessing the current method of screening for mortality and morbidity conferences.
 
@@
 
-Trauma, defined as a physical injury and the body’s associated response, causes 8-9% of global deaths annually [@world2014injuries;
-@roth2018global]and ranks highest in disability-adjusted life-years (DALYs) [@vos2020global],
-impacting individuals, communities and society at large
-[@newnam2014impacts]. Road traffic injuries alone cost approximately 2%
-of gross domestic product (GDP) in high-income countries
-[@world2014injuries].
+Trauma, defined as a physical injury and the body's associated response,
+causes 8-9% of global deaths annually [@world2014injuries;
+@roth2018global] and ranks highest in disability-adjusted life-years
+(DALYs) between ages 10-49 years [@vos2020global], impacting
+individuals, communities and society at large [@newnam2014impacts; @david2022measuring]. Road
+traffic injuries alone cost approximately 2% of gross domestic product
+(GDP) in high-income countries [@world2014injuries]. Despite
+advancements in trauma care, preventable deaths [@kwon2014pooled;
+@ghorbani2018analysis; @teixeira2007preventable; @vioque2014classifying] and errors persist [@nikouline2021errors;
+@vioque2014classifying; @o2013opportunities; @teixeira2007preventable]. To address this, trauma quality
+improvement programs [@American2022resources], commonly including
+mortality and morbidity conferences [@world2009guidelines] have been
+implemented. The conferences have shown to foster error reduction and
+encourage reporting [@kong2016analysis] without negative feedback
+[@lazzara2022morbidity].
 
-Quality improvement programs form the core of current trauma care systems 
-[@American2022resources] and commonly include
- techniques such as multidisciplinary mortality and morbidity conferences and preventable death review panels to improve trauma care quality
-[@world2009guidelines]. Despite these programs, trauma care remains prone to errors [@nikouline2021errors; @vioque2014classifying;
-@o2013opportunities].
-Studies have shown a pooled preventable death rate of 20%, with more recent studies indicating lower rates [@kwon2014pooled].
-A recent study in a Swedish Level I trauma centre found a preventable death rate of 4%
-[@ghorbani2018analysis].
+Audit filters are criteria used to identify cases for mortality
+and morbidity conferences aiming to pinpoint potential opportunities for
+improvement [@American2022resources]. Audit filters can involve outcomes
+such as "death" or adherence to guidelines, such as "delay to treatment"
+[@world2009guidelines]. Opportunities for improvement are typically
+associated with failures in initial care [@vioque2014classifying] like
+airway management, fluid resuscitation, haemorrhage control and chest
+injury management [@sanddal2011analysis; @roy2017learning;
+@o2013opportunities]. There seems to be a lack of common, evidence-based
+set of audit filters for evaluating trauma care quality
+[@stelfox2010quality], and while early studies reported mortality
+reduction [@chadbunchachai2001comparative; @chadbunchachai2003study],
+newer studies differ [@evans2009audit; @willis2008interpreting]. Moreover, the use of audit filters has been
+associated with high frequencies of false positives, ranging from 24% to
+80% in different contexts [@sanddal2011analysis; @roy2017learning; @ghorbani2018analysis]. Few
+studies have focused on
+non-mortality outcome measures, for evaluating performance of audit
+filters but the evidence remains inconsistent [@di2008acscot;
+@glance2012association] and
+some audit filters do not appear to correlate with opportunities for
+improvement at all [@lewis2017delay]. However, the perceived
+usefulness of audit filters remains high [@berg2022perceived]. The aim of this study is to determine the performance of individual
+audit filters in predicting
+opportunities for improvement and their relevance as a
+screening tool for mortality and morbidity conferences
 
-The mortality and morbidity conferences is a central technique focusing on anonymity, critical analysis of specific adverse events and recognizing flawed approaches followed by proposing and implementing improvements [@gregor2016morbidity].
-Attended by healthcare professionals involved in trauma care, mortality and morbidity conferences discuss themes such as unexpected mortality, morbidity and errors
-[@orlander2003morbidity], fostering error reduction and encouraging
-reporting without negative feedback [@lazzara2022morbidity].
-Effective mortality and morbidity conferences implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].
-
-Audit filters are used to identify cases for mortality and morbidity conferences.
-These filters are criteria based on standards of care, guidelines or outcomes and function as a screening tool for potential opportunities for improvement.
-Deviation from audit filters should raise the attention of health care providers with the purpose of improving the quality of care
-[@American2022resources]. Audit filters can involve outcomes such as “death” or adherence to guidelines, such as “placement of two large bore intravenous lines within 15 minutes of arrival.”
-[@world2009guidelines].Studies have shown conflicting results on the effectiveness of audit filters in reducing trauma-related mortality.
-While some early studies reported a mortality reduction following the implementation of audit filters
-[@chadbunchachai2001comparative; @chadbunchachai2003study], a 2009
-review by Evans et al. [@evans2009audit]  found no high quality evidence supporting the effectiveness of audit filters. Furthermore, Stelfox et al.  [@stelfox2010quality]
-identified over 1,500 audit filters and noted a lack of a common, evidence-based set for evaluating trauma care quality.
-Despite the uncertainty, the perceived usefulness of audit filters remains high [@berg2022perceived].
-
-Opportunities for improvement are decided upon review of individual patient cases and aim to reflect current trauma care quality gaps  [@American2022resources].
-These opportunities are typically associated with failures in initial care [@vioque2014classifying]  like airway management, fluid resuscitation, haemorrhage control and chest injury management [@sanddal2011analysis; @roy2017learning;
-@o2013opportunities]. There is lacking evidence for the use of audit filters for improving trauma care [@evans2009audit; @stelfox2010quality], but the focus has been on mortality as outcome measure 
-[@chadbunchachai2001comparative; @chadbunchachai2003study].Moreover, the use of audit filters  has been associated with high frequencies of false positives, ranging from 24% to 80% [@sanddal2011analysis;
-@roy2017learning], and some audit filters do not appear to correlate with opportunities for improvement at all [@lewis2017delay].
-
-The aim of this study is to determine the performance of audit filters in predicting opportunities for improvement and assessing their relevance as a screening tool for mortality and morbidity conferences
-
-```{=html}
-<!-- Explain the scientific background and rationale for the
-investigation being reported. End by stating the aim or specific
-objectives, including any prespecified hypotheses. The introduction
-should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
-6-8 lines long. Support every statement with a citation -->
-```
 # Methods
@@
 
-We conducted a single-centre registry-based retrospective cohort study  of all trauma patients included in both the Karolinska University Hospital trauma registry and trauma care quality database between 2012 and 2022 to compare the performance of audit filters in their ability to predict opportunities for improvement.
-This study was approved by Swedish Ethical Review Authority, approval number 2021-02541 and 2021-03531.
+We conducted a single-centre registry-based retrospective cohort study
+of all trauma patients included in both the Karolinska University
+Hospital trauma registry and trauma care quality database between 2013
+and 2022 to compare the performance of audit filters in their ability to
+predict opportunities for improvement. This study was approved by the
+Swedish Ethical Review Authority, approval number 2021-02541 and
+2021-03531.
 
@@
 
-Karolinska University Hospital manages around 1,500 trauma patients annually.
-The Karolinska Trauma Registry, which reports to the national trauma registry SweTrau, includes all patient cases that result in trauma team activation or who have an Injury Severity Score (ISS) of more than 9.
-This registry includes data on vital signs, timings, injuries, interventions, and patient demographics as per a European consensus statement
-[@ringdal2008utstein]
+Karolinska University Hospital manages around 1,500 trauma patients
+annually. The Karolinska Trauma Registry, which reports to the national
+trauma registry SweTrau [@SweTrau], includes all patients that result in
+trauma team activation or who have an Injury Severity Score (ISS) of
+more than 9. This registry includes data on vital signs, timings,
+injuries, interventions, and patient demographics as per the European
+consensus statement [@ringdal2008utstein]
 
-The trauma centre at the Karolinska University Hospital also keeps a local care quality database with information intended for the mortality and morbidity conferences, including identified triggered audit filters and opportunities for improvement.
-Opportunities for improvement are assessed in multiple stages of scrutiny.
-Notably, mortality is directly referred to the mortality and morbidity conferences to assess opportunities for improvement.
+The trauma centre at the Karolinska University Hospital also keeps a
+local care quality database with information intended for the mortality
+and morbidity conferences, including identified triggered audit filters
+and opportunities for improvement. Opportunities for improvement are
+assessed in multiple stages of scrutiny. Notably, mortality is directly
+referred to the mortality and morbidity conferences to assess
+opportunities for improvement.
 
-From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality.
-Initially, trauma cases were individually assessed by a specialized trauma nurse for potential opportunities for improvement.
-In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses.
-Cases flagged by an audit filter or identified by the nurse as potential care failures underwent a second review by two specialized nurses.
-Potential opportunities for improvement identified in this review prompted review by the mortality and morbidity conference.
+During the study period, the screening of patient cases for potential
+opportunities for improvement was formalised. Prior to 2017, not all
+patient cases were subjected to an initial evaluation by a specialised
+nurse. However, following 2017 all patient cases underwent screening.
+Importantly, audit filters and an initial evaluation by a specialised
+nurse has consistently functioned as the screening tool for mortality
+and morbidity conferences to identify potential opportunities for
+improvement.
 
-```{=html}
-<!-- Describe the setting, locations, and relevant dates, including
-periods of recruitment, exposure, follow-up, and data collection -->
-```
 ## Participants
 
-The study included all patients treated at Karolinska University Hospital from 2012 to 2022 who underwent screening for opportunities for improvement.
-Patients under 15 were excluded due to differences in clinical management.
+The study included all patients treated at Karolinska University
+Hospital from 2013 to 2022 who underwent screening for opportunities for
+improvement. Patients aged under 15 years were excluded due to differences in
+clinical management.
 
-```{=html}
-<!-- Cohort study: Give the eligibility criteria, and the sources and
-methods of selection of participants. Describe methods of
-follow-up. For matched studies, give matching criteria and number of
-exposed and unexposed -->
-```
-```{=html}
-<!-- Case-control study: Give the eligibility criteria, and the
-sources and + methods of case ascertainment and control
-selection. Give the rationale for the choice of cases and
-controls. For matched studies, give matching criteria and the number
-of controls per case -->
-```
-```{=html}
-<!-- Cross-sectional study: Give the eligibility criteria, and the
-sources and methods of selection of participants -->
-```
 ## Variables and data sources/measurements
-The outcome was opportunities for improvement, as established by the mortality and morbidity conference, and was treated dichotomously as “Yes - At least one opportunity for improvement identified” and “No - No opportunity for improvement identified”.
 
-The eleven audit filters used at the Karolinska University Hospital served as the exposure variables, as shown in Table 1.
-The data spanning 2012 to 2022 was retrieved from the Karolinska Trauma Registry and trauma care quality database.
-Patient data on vital signs, care processes and interventions, level of care and time aspects was retrieved from SweTrau while both the exposure audit filters and the outcome opportunities for improvement were retrieved from the quality database.
+The outcome was opportunities for improvement, as established by the
+mortality and morbidity conference, and was treated dichotomously as
+"Yes - At least one opportunity for improvement identified" and "No - No
+opportunity for improvement identified".
 
-All data was anonymised to protect patient privacy.
+The ten audit filters used at the Karolinska University Hospital served
+as the exposure variables.
+For each audit filter, we identified the cohort of patients in which the filter was applicable, and evaluated the filters’ performance in that cohort: For “ISS > 15 and no team activation” and “ISS > 15 and not in ICU” we evaluated the filters’ performance in patients with an ISS greater than 15; for "GCS < 9 and not intubated" we analyzed patients with a GCS less than 9; and lastly, for "No anticoagulants within 72 hours after TBI" we identified patients with head trauma as defined by the precence of at least one head injury according to available AIS codes.
 
-```{=html}
-<!-- Clearly define all outcomes, exposures, predictors, potential
-confounders, and effect modifiers. Give diagnostic criteria, if
-applicable. For each variable of interest, give sources of data and
-details of methods of assessment (measurement). Describe comparability
-of assessment methods if there is more than one group -->
-```
-## Bias
+The data was retrieved from the Karolinska Trauma Registry and trauma
+care quality database. Patient data on vital signs, care processes and
+interventions, level of care and time aspects was retrieved from the trauma registry whereas the exposure, audit filters and the outcome, and opportunities
+for improvement were retrieved from the quality database.
 
-Consensus decisions at the mortality and morbidity conferences may introduce misclassification bias.
-Additionally, audit filters flagging by specialized nurses is a manual process, potentially leading to selection bias.
-Healthcare professionals attending mortality and morbidity conferences may not accurately recall key details impacting opportunities for improvement decisions, introducing recall bias.
-Furthermore, professionals, not present for specific cases might over- or underestimate opportunities for improvement significance.
-```{=html}
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
-```
 ## Statistical methods
 
- for statistical analysis.
-Patient characteristics for cases with and without opportunities for improvement were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.
-
-The predictive performance of audit filters was assessed by calculating the sensitivity, specificity and Cohen’s kappa for each audit filter in predicting opportunities for improvement.
-Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and Cohen’s kappa coefficient.
-
-Missing values for audit filters were treated as negative (i.e., “not flagged”).
+We used R [@R2022language] for statistical analysis. Patient
+characteristics for cases with and without opportunities for improvement
+were compared using the Wilcoxon signed-rank test for continuous
+variables and the Chi-squared test for categorical variables. The
+predictive performance of audit filters was assessed by calculating the
+sensitivity, specificity and Cohen's kappa for each audit filter in
+predicting opportunities for improvement. Bootstrapping with 1,000
+iterations was applied to estimate 95% confidence intervals for all
+performance metrics. We selected cohorts relevant to each filter: For
+"ISS \> 15 and no team activation" and "ISS \> 15 and not in ICU" we
+examined patients with an ISS greater than 15. For "GCS \< 9 and not
+intubated," we focused on patients with a GCS less than 9. Lastly, for
+"No anticoagulants within 72 hours after TBI" we identified patients
+with head trauma as defined by the presence of at least one head injury
+according to available AIS codes. Missing values were numbered but
+ignored in the analysis.
 
@@
 
-##Participants 
-After excluding patients under 15 years old and patients not screened for opportunities for improvement, a total of 8,309 individuals were included in the study.
-7,797 out of 8,309 (93.8%) patient cases did not have an opportunity for improvement and 512 out of 8,309 (6.2%) had an opportunity for improvement, see Figure 1.
+After excluding patients aged less than 15 years and patients not screened
+for opportunities for improvement, a total of 8,309 individuals were
+included in the study. Out of these 8,309 patients, a total of 512 (6.2%) patients had an
+opportunity for improvement, see Figure 1.
 
-Patients with an opportunity for improvement had significantly higher mean age `r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs `r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
-`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`, ISS `r inline_text(tableOne, variable = ISS, column = stat_2)` vs
-`r inline_text(tableOne, variable = ISS, column = stat_1)`, `r inline_text(tableOne, variable = ISS, column = "p.value")` and intubations in the emergency department (ED) `r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
-vs `r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)`,
-`r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
-Patients with opportunity for improvement had longer times to definitive treatment from hospital arrival `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)` vs
-`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`,
+Patients with an opportunity for improvement had significantly higher
+median age,
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` years vs
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)` years,
+`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`, median 
+ISS `r inline_text(tableOne, variable = ISS, column = stat_2)` vs
+`r inline_text(tableOne, variable = ISS, column = stat_1)`,
+`r inline_text(tableOne, variable = ISS, column = "p.value")` and
+intubations in the emergency department (ED), 
+`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)` patients
+vs
+`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)` patients,
+`r inline_text(tableOne, variable = ed_intubated, column = "p.value")`.
+Patients with opportunity for improvement had longer times to first major intervention from hospital arrival median
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
+minutes vs
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)` minutes,
 `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
- and longer times to first CT
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`,
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`. The number of opportunities for improvement were highest in patients admitted to the intensive care unit 
-`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)` compared to general ward `r inline_text(tableOne, variable = host_care_level, level = "General ward", column = stat_2)`.
-Gender, mean systolic blood pressure, death at 30 days and trauma team activation was not significantly correlated with opportunities for improvement (p>0.05).
-Notably, there was a statistically significant correlation between GCS at ED and opportunities for improvement (p < 0.001).
-The highest amount of missing data was observed for the audit filter "time to
-definitive treatment" (n = 5,990, 72.1%). The smallest amount of missing data was found in "SBP\<90" (n = 13, 0.2%).
+and longer times to first CT, median
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` minutes vs
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)` minutes,
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")` minutes.
+The frequency of opportunities for improvement was highest in patients
+admitted to the intensive care unit,
+`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)`
+patients, compared to general ward
+`r inline_text(tableOne, variable = host_care_level, level = "General ward", column = stat_2)`. Notably, there was a statistically significant
+correlation between GCS at ED and opportunities for improvement (p \<
+0.001).
+Gender, mean systolic blood pressure, death at 30 days and trauma team
+activation were not significantly correlated with opportunities for
+improvement (p\>0.05). The highest amount of missing data was observed for the audit
+filter "time to first major intervention", which was missing for 5,990 patients, (72.1%). See Table 2
+for details of selected patient characteristics.
 
@@
 
-The number of patient cases flagged by each individual audit filter varied between 107 for “GCS<9 and not intubated” and 3907 for “>30 min until CT”.
-The highest number of missing values was observed with “>60 min until first intervention” (n = 5990, 72.1%).
-The lowest number of missing values was observed with “dead at 30 days” (n = 11, 0.1%) and “ISS > 15 and not in ICU” (n = 11, 0.1%).
-See table 2 for missing values for each audit filter.
+The number of patient cases flagged by each individual audit filter
+varied between (n = 107, 27.4%) for "GCS \< 9 and not intubated" and (n
+= 3907, 53.5%) for "\>30 min until CT". The highest number of missing
+values was observed with "\> 60 min until first intervention" (n = 5990,
+72,1%). The lowest number of missing values was observed with "dead at
+30 days" and "ISS \> 15 and not in ICU" (n = 11, 0.1%). See table 1 for
+missing values for each audit filter. The largest cohort was observed
+with "Dead at 30 days" (n = 8298, 99.9%) and the smallest cohort was
+observed with "GCS \< 9 and not intubated" (n = 390, 4.7%). In total
+5977 out of 8309 (71.9%) patient cases were flagged by at least one
+audit filter.
 
-The audit filter with the highest sensitivity was “>30 min until CT” (59.2%, 95% CI [55.2-63.8]).
-The lowest sensitivity was seen in “GCS<9 and not intubated” (1.2%, 95% CI [0.2-2]).
-The audit filter with the highest specificity was “GCS<9 and not intubated” (98.7%, 95% CI [98.5-98.9]).
-The lowest specificity was seen in “>30 min until CT” (53.8%, 95% CI [52.7-54.8]).
-Notably, the confidence intervals varied between sensitivity and specificity, with sensitivity displaying  wider intervals.
-See Table 4 for the performance of all audit filters using sensitivity and specificity.
+The audit filter with the highest sensitivity was "\> 60 min until first
+intervention" (83.3%, 95% CI [79.2-87.5]). The lowest sensitivity was
+seen in "SBP \< 90" (8%, 95% CI [5.6-10.2]) and "Dead at 30 days" (8%,
+95% CI [5.6-10.2]). The audit filter with the highest specificity was
+"Massive transfusion" (95.8%, 95% CI [95.4-96.3]). The lowest
+specificity was seen in "ISS \> 15 and no team activation" (21.6%, 95%
+CI [19.9-23.2]). The sensitivity for all audit filters combined was
+92.6% (95% CI [90.5-95]) and the specificity was 29.4% (95% CI
+[28.4-30.5]).
 
-Two of the eleven audit filters, “GCS < 9 and not intubated” and “CPR and thoracotomy”, had a Cohen’s kappa coefficient of 0.
-The highest Cohen’s kappa was seen for the audit filter “>60 min until first intervention” and “ISS > 15 and not in ICU” (0.12, 95% CI [0.10-0.15]).
-All audit filters showed none to slight correlation with opportunities for improvement.
-See table 5 for the performance of all audit filters using Cohen’s kappa coefficient.
+The highest Cohen's kappa was seen for the audit filter "Massive
+transfusion" (0.08, 95% CI [0.05-0.12]). The smallest Cohen's kappa was
+seen in "GCS \< 9 and not intubated" (-0.07, 95% CI [-0.14-0]. The
+Cohen´s kappa for all audit filters combined was 0.04 (95% CI
+[0.03-0.04]. See table 3 for the performance of all audit filters using
+sensitivity, specificity and Cohen's kappa coefficient.
 
@@
 
-```{=html}
-<!--
-
-Participants
-------------
-
-(a) Report numbers of individuals at each stage of study—eg numbers
-potentially eligible, examined for eligibility, confirmed eligible,
-included in the study, completing follow-up, and analysed
-
-(b) Give reasons for non-participation at each stage
-
-(c) Consider use of a flow diagram
-
-Descriptive data
-----------------
-
-(a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
-
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
-
-Outcome data
-------------
-
-Cohort study — Report numbers of outcome events or summary measures
-over time
-
-Case-control study — Report numbers in each exposure category, or
-summary measures of exposure
-
-Cross-sectional study — Report numbers of outcome events or summary
-measures
-
-Main results
-------------
-
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
-
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
-
-Other analyses 
---------------
-
-Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses 
--->
-```
-You can include code in this document like this:
-
-```{r main, echo=FALSE}
-source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
-
-You can also embed plots:
-
-```{r plot, echo=FALSE}
-plot(pressure)
-```
-
-You can also mix text and code, so called inline code, like this:
-`r 2+5`.
-
 # Discussion
 
-The aim of this study was to evaluate the performance of individual audit filters in predicting opportunities for improvement and to assess their suitability as a screening tool for mortality and morbidity conferences.
-This study showed that audit filters had “no” to “slight” ability to predict opportunities for improvement.
-Two audit filters, “GCS<9 and not intubated” and “CPR and thoracotomy”, showed no predictive ability.
-The highest Cohen’s kappa coefficient was seen for the audit filter “>60 min until first intervention” and “ISS > 15 and not in ICU”, but it was still low.
+This study  shows that audit filters, individually or in combination,  were not effective in predicting opportunities for improvement.
+As a result, the development of a new screening method to better identify potential opportunities for improvement is advised.
 
-## Limitations
+agree
 
-Opportunity for improvement, while binarily defined, includes a diverse set of outcomes.
-This complicates the process of creating audit filters that perform well in predicting opportunities for improvement.
-Audit filters would favour the identification of some, but not all, errors.
+disagree
+Some previous studies have shown similar findings to this study.
+Lewis et al.
+looked at one audit filter, "delay to operating room" of more than 2 hours and its correlation with opportunities for improvement.
+They found that the audit filter had no significant ability to predict opportunities for improvement.
+Similarly, the delay-related audit filters: "\>60 min until first intervention" and
+"\>30min until first CT" showed noability to predict opportunities for improvement.
+A big meta analysis by Evans et al.
+sought to identify a common set of evidence based audit filters but no study fulfilled the inclusion criteria.
+A 2008 study by Di Bartolomeo et al.
+[@di2008acscot] looked at 4 ASCOT audit filters in relation to 30-day mortality and a morbidity measure.
+One audit filter showed a correlation with mortality but no audit filter showed a correlation with the morbidity measure.
 
-Each audit filter was individually related to opportunities for improvement.
-Consequently, if an audit filter demonstrated a true positive correlation with an opportunity for improvement, instances where the same patient case had multiple audit filters introduce uncertainty regarding which specific audit filter truthfully predicted the opportunity for improvement.
+A recent study by Berg et al. [@berg2024effects], looking at implementation of audit filters on mortality concluded that audit filters might be useful in reducing the mortality. 
+Previous studies suggest poor performance of audit filters for improving
+trauma care [@evans2009audit; @stelfox2010quality; @willis2008interpreting; @di2008acscot; @glance2012association; @lewis2017delay].
+This study adds to that evidence by showing that audit filters perform poorly in predicting opportunities for
+improvement encompassing mortality and non-mortality
+associated outcomes. Some audit filters might be more reflective of clinical decisions rather than quality of care [@glance2012association]. For instance, an abnormal GCS because of intoxication
+is managed differently to an abnormal GCS due to head injury, which could explain the poor performance of "GCS < 9 and not intubated", similarly shown by
+Willis et al. [@willis2008interpreting]. This could also explain the poor performance of the delay-related audit filters: "\>60 min until first intervention" and
+"\>30min until first CT". Similarly, Lewis et al. [@lewis2017delay] showed that delay
+to the operating room was not associated with opportunities for improvement.
+In contrast, Teixeira et al. [@teixeira2007preventable] identified delays to
+treatment and delays to CT as common errors in trauma care. A recent
+study in Japan by Yamamoto et. al [@yamamoto2023immediate] observed
+significantly reduced mortality with whole-body CT scans within 10
+minutes of arrival to the ED.
+If anything, this highlights the unique opportunities for improvement in different hospitals.
+The audit filters also varied greatly in terms of expected mortality.
+Some audit
+filters such as "Massive transfusion" and "Dead at 30 days" have a high
+expected mortality. In such cases the reduced time frame for any potential
+opportunity for improvement to occur could explain their poor performance.
+When audit filters were combined they had a low specificity relative to its sensitivity which corresponds to previous studies showing high numbers of false positives [@sanddal2011analysis; @roy2017learning; @ghorbani2018analysis].
+The sensitivity and specificity for individual audit filters varied substantially, suggesting that some audit filters are too rigid, missing the majority of opportunities for improvement while some are too inclusive, causing high numbers of false positives [@willis2008interpreting].
 
-The number of missing values varied greatly among the audit filters.
-As stated above, missing values from each audit filter were coded as “not flagged” by an audit filter.
-This means that instances where patient cases were not flagged might meet the audit filter criteria, but were not documented.
-This reduces the validity of audit filters that feature high amounts of missing data.
+There were some limitations in this study.
+Firstly, it is unlikely that the specific associations presented in this study are generalizable to other hospitals with similar systems, but the broader finding that audit filters are poorly associated with opportunities for improvement is likely to be externally valid [@world2009guidelines].
+Secondly, instances where the
+same patient case had multiple audit filters introduce uncertainty
+regarding which specific audit filter truthfully predicted an
+opportunity for improvement. Finally, the number of missing values
+differed significantly among the audit filters, resulting in varying
+degrees of validity for each filter.
 
-##Interpretation
-Higher ISS, longer times to CT, longer times to first intervention, and higher levels of care correlated with opportunities for improvement, consistent with a recent Swedish study by Albaaj et al.[@albaaj2023patient].
-Notably, GCS at the ED also showed a correlation with opportunities for improvement despite identical median GCS values, likely due to distribution differences outside the interquartile range (IQR).
-
-The audit filter “GCS<9 and not intubated” showed no significant ability to predict opportunities for improvement, similarly shown by Willis et al. 
-[@willis2008interpreting]. Possibly because the audit filter might be more reflective of clinical decisions rather than quality of care 
-[@willis2008interpreting]. For instance, an abnormal GCS because of intoxication is managed differently to an abnormal GCS due to head injury.
-
-Both delay-related audit filters: “>60 min until first intervention” and “>30min until first CT”, showed a slight correlation with opportunity for improvement.
-Comparably, Teixeira et al.[@teixeira2007preventable] linked delays to preventable death andidentified delay to treatment and delays to CT as common errors in trauma care.
-A recent study in Japan by Yamamoto et.
-al [@yamamoto2023immediate] observed significantly reduced mortality with whole-body CT scans within 10 minutes of arrival to the ED.
-
-“CPR and thoracotomy” showed no correlation with opportunities for improvement.
-This could be explained by non-preventable mortality, thus presenting a smaller time frame for mistakes to be made.
-A similar explanation could be given to the inverse relation between “Dead at 30 days” and opportunities for improvement.
-
-Audit filters generally exhibited a high specificity relative to their sensitivity which may be attributed to the outcome being rather uncommon, whilst every individual audit filter misses the majority of opportunities for improvement.
-This could suggest that the audit filters are too rigid in their criteria.
-On the contrary, too inclusive criteria might flag several patients, without opportunities for improvement [@willis2008interpreting].
-
-## Generalizability
-
-Audit filters are widely used, but lack consensus on which to use, purpose, and application [@stelfox2010quality; @stelfox2011evidence;
-@berg2022perceived].
-They have static criteria and may therefore not adapt to evolving healthcare challenges, or upon resolution of the quality gaps they were designed for, reducing their utility over time.
-This underscores the need for regular evidence-based reviews of audit filter performance [@santana2014development].
-However, these reviews might be cost-prohibitive.
-
-While audit filters can help identify an opportunity for improvement, they might overshadow other contributing factors, leading to over-reliance and potentially failing to recognize cases outside their scope.
-[@willis2008interpreting].For example, audit filters often focus on aspects relevant to doctors, despite the interdisciplinary nature of trauma care.
-Lastly, confounding variables, such as different treatment times based on injury severity, add to the uncertainty of audit filters’ effectiveness.
+The static nature of audit filters inherently limit their adaptability to
+emerging healthcare challenges and resolved quality gaps, reducing their
+utility over time. Therefore, regular evidence-based reviews are needed
+to ensure performance [@santana2014development], though such reviews may
+be cost-prohibitive. Furthermore, they might overshadow other
+contributing factors to opportunities for improvement
 [@willis2008interpreting].
 
-Due to the study being conducted in a single trauma centre in Stockholm it is unrealistic to assume applicability of these findings at other hospitals facing unique challenges [@world2009guidelines].
-For instance, audit filters that benefit low and middle income may not benefit high income countries and vice versa [@berg2022perceived].
-
-## Future Studies
-
-# Conclusion
-
-Audit filters are widely used as a tool in quality improvement programmes and should reflect the current gap in trauma care quality.
-In this study, no audit filter showed a significant performance in predicting opportunities for improvement, with Cohen’s kappa coefficient ranging from -0.01 to 0.12.
-These results highlight the importance of continually evaluating the relevance of audit filters.
+In summary, audit filters are widely used as a tool in quality
+improvement programmes and should reflect current gaps in trauma care
+quality. In this study, no individual audit filter or their combination were
+successful in predicting opportunities for improvement. Given these
+results, future studies should focus on developing new tools with higher predictive
+ability.
 
 # Acknowledgements
-The authors thank all professionals taking part in the monthly mortality and morbidity conferences.
+
+The authors thank all professionals taking part in the monthly mortality
+and morbidity conferences.
 

### 2024-09-14 functions/confidence_interval_sens_spec.R

@@ -1,8 +1,8 @@
 confidence_interval_sens_spec <- function(twoVariableData){
-  set.seed(123) #Setting the seed for bootstrap
+  set.seed(122) #Setting the seed for bootstrap
   numberOfBootstraps <- 1000 #Determining number of bootstrap iterations
   bootResults <- boot(twoVariableData,statistic = bootstrap_sens_spec, R = numberOfBootstraps) #bootstrapping 
-  sensitivityCI <- boot.ci(bootResults, type = "basic", index = 1)  # Sensitivity CI
+  sensitivityCI <- boot.ci(bootResults, type = "basic", index = 1) # Sensitivity CI
   specificityCI <- boot.ci(bootResults, type = "basic", index = 2) #calculating confidence interval from bootstrap
   return(c(round(sensitivityCI$t0 * 100, digits = 1), round(sensitivityCI$basic[4:5] * 100, digits = 1), round(specificityCI$t0 * 100, digits = 1), round(specificityCI$basic[4:5] * 100, digits = 1))) # returning the original cohen kappa followed by the confidence interval
 }

### 2024-09-14 functions/create_audit_filters.R

@@ -1,17 +1,15 @@
 create_audit_filters <- function(cleanData){
-
+### SAP < 90 ###
   cleanData$AF_sap_less90 <- ifelse(
     is.na(cleanData$ed_sbp_value) & is.na(cleanData$ed_sbp_rtscat), NA, 
     ifelse(is.na(cleanData$ed_sbp_value), cleanData$ed_sbp_rtscat <= 3, 
            ifelse(cleanData$ed_sbp_value < 90 , TRUE,FALSE)))
-  #cleanData$AF_sap_less90 <- cleanData$ed_sbp_value < 90 | cleanData$ed_sbp_rtscat <= 3
-  #cleanData$AF_sap_less90[is.na(cleanData$AF_sap_less90)] <- FALSE
   #VK_sap_less90
+  cleanData$GCS_under_9 <- cleanData$ed_gcs_sum < 9 & !is.na(cleanData$ed_intubated)
+  cleanData$GCS_under_9_M <- is.na(cleanData$ed_gcs_sum) | is.na(cleanData$ed_intubated)
   cleanData$AF_gcs_less9_ej_intubTE <- ifelse(
     is.na(cleanData$ed_gcs_sum) | is.na(cleanData$ed_intubated) | cleanData$ed_gcs_sum >= 9, NA, 
     ifelse(cleanData$ed_gcs_sum < 9 & cleanData$ed_intubated != "Yes", TRUE,FALSE))
-  #cleanData$AF_gcs_less9_ej_intubTE <- cleanData$ed_gcs_sum < 9 & cleanData$ed_intubated != "Yes"
-  #cleanData$AF_gcs_less9_ej_intubTE[is.na(cleanData$AF_gcs_less9_ej_intubTE)] <- FALSE
   #VK_gcs_less9_ej_intubTE
   cleanData$AF_mer_30min_DT <- cleanData$dt_ed_first_ct > 30
 
@@ -21,23 +19,25 @@ create_audit_filters <- function(cleanData){
   #cleanData$AF_mer_60_min_interv[is.na(cleanData$AF_mer_60_min_interv)] <- FALSE
   #VK_mer_60_min_interv
   #cleanData$AF_iss_15_ej_iva <- cleanData$ISS >= 15 & cleanData$host_care_level != "Intensive care unit"
+  cleanData$ISS_over_15 <- cleanData$ISS >= 15 & !is.na(cleanData$host_care_level)
[…]

### 2024-09-14 functions/select_data.R

@@ -3,7 +3,7 @@ select_data <- function(combinedDataset) {
   selectedData <- select(combinedDataset,starts_with("AF_"), ofi,
                          Gender,ISS, pt_age_yrs, ed_sbp_value, ed_gcs_sum,
                          dt_ed_first_ct, ed_intubated,res_survival,dt_ed_emerg_proc,
-                         host_care_level,Tr_Nivå)
+                         host_care_level,Tr_Nivå, TBI, GCS_under_9, ISS_over_15, ISS_over_15_T, TBI_M, GCS_under_9_M, ISS_over_15_M, ISS_over_15_T_M)
   #select what i want
   return(selectedData)
 }

### 2024-09-14 main.R

@@ -1,12 +1,12 @@
 library(rofi)
 library(dplyr)
 library(ggplot2)
-library(pROC)
 library(gtsummary)
 library(gt)
 library(boot)
 library(magrittr)
-library(Gmisc, quietly = TRUE)
+library(knitr)
+#library(Gmisc, quietly = TRUE)
 library(glue)
 library(htmlTable)
 library(grid)
@@ -17,37 +17,27 @@ library(DiagrammeR)
 listOfAuditFilters <- c("AF_sap_less90","AF_death_30d","AF_iss_15_ej_TE",
                         "AF_mass_transf", "AF_gcs_less9_ej_intubTE", 
                         "AF_iss_15_ej_iva", "AF_mer_60_min_interv",
-                        "AF_mer_30min_DT", "AF_hlr_thorak", 
+                        "AF_mer_30min_DT", 
                         "AF_lever_och_mjaltskada", "AF_ej_trombrof_TBI_72h", "AF_all")
 listOfAuditFiltersClean <- c("SBP < 90", "Dead at 30 days", "ISS > 15 and no team activation", 
                              "Massive transfusion", "GCS < 9 and not intubated", "ISS > 15 and not in ICU", 
-                             "> 60 min until first intervention", "> 30 min until first CT",
-                             "CPR and thoracotomy", "Liver or spleen injury", 
+                             "> 60 min until first intervention", "> 30 min until first CT", "Liver or spleen injury", 
                              "No anticoagulants within 72 hours after TBI", "All")
-
-#1,2,3,5,6,7,8 manual
 ## viktigt att det ska vara samma ordning ^
-selectedAuditFilter <- listOfAuditFilters[1:12]
+DefOfAbbreviations <- "Definition of abbreviations: SBP = Systolic Blood Pressure; ISS = Injury Severity Score; GCS = Glascow Coma Scale; ICU = Intensive Care Unit; CT = Computer Tomography; ED = Emergency Department; CPR = Cardiopulmonary Resuscitation; TBI = Traumatic Brain Injury"
+
+selectedAuditFilter <- listOfAuditFilters[1:11]
 
 tableKappa <- data.frame(Auditfilter = character(0),
                          Sensitivity = numeric(0),
                          Specificity = numeric(0),
                          Kappa = numeric(0))
 tableAuditFilter <- data.frame(Auditfilter = character(0),
[…]

### 2024-10-07 manuscript.Rmd
update manuscript

@@
 library(webshot2)
+
 ```
@@
 Performance was measured using sensitivity, specificity and Cohen’s kappa coefficient for ten audit filters.
-Bootstrapping with 1,000 iterations calculated 95% confidence intervals.
 
@@
 Our analysis included 8,309 patients.
-Audit filters showed a Cohen’s kappa coefficient ranging from -0.07 (95% CI [-0.14-0]) for “GCS < 9 and not intubated” and 0.08 (95% CI [0.05-0.12]) for “Massive transfusion”.
-The Cohen’s kappa coefficient for all audit filters combined was 0.04 (95% CI[0.03-0.04]).
+Audit filters demonstrated inconsistent agreement between filters and opportunities for improvement with a Cohen’s kappa coefficient ranging from -0.07 (95% confidence interval [CI], -0.14-0) for “GCS < 9 and not intubated” and 0.08 (CI, 0.05-0.12) for “Massive transfusion”.
+The Cohen’s kappa coefficient for all audit filters combined was 0.04 (CI, 0.03-0.04).
 
@@
 
-No individual audit filter performed well in predicting opportunities for improvement, which suggests reassessing the current method of screening for mortality and morbidity conferences.
+No individual audit filter performed well in predicting opportunities for improvement, which suggests that the current method of screening for mortality and morbidity conferences should be revisited.
+
+Level III, Prognostic/Epidemiological
+ofi, audit filter, quality improvement
 
@@
 advancements in trauma care, preventable deaths [@kwon2014pooled;
-@ghorbani2018analysis; @teixeira2007preventable; @vioque2014classifying] and errors persist [@nikouline2021errors;
+@ghorbani2018analysis; @teixeira2007preventable; @vioque2014classifying] and avoidable medical errors persist [@nikouline2021errors;
 @vioque2014classifying; @o2013opportunities; @teixeira2007preventable]. To address this, trauma quality
@@
 injury management [@sanddal2011analysis; @roy2017learning;
-@o2013opportunities]. There seems to be a lack of common, evidence-based
+@o2013opportunities]. There is a lack of common, evidence-based
 set of audit filters for evaluating trauma care quality
@@
 and opportunities for improvement. Opportunities for improvement are
-assessed in multiple stages of scrutiny. Notably, mortality is directly
-referred to the mortality and morbidity conferences to assess
+assessed in multiple stages of scrutiny. Notably, patient deaths are always reviewed to determine if they could have been prevented and to identify
 opportunities for improvement.
@@
 The study included all patients treated at Karolinska University
-Hospital from 2013 to 2022 who underwent screening for opportunities for
-improvement. Patients aged under 15 years were excluded due to differences in
-clinical management.
+Hospital from 2013 to 2022 and who were screened for opportunities for
+improvement. Patients aged under 15 years were excluded.
 
@@
 as the exposure variables.
-For each audit filter, we identified the cohort of patients in which the filter was applicable, and evaluated the filters’ performance in that cohort: For “ISS > 15 and no team activation” and “ISS > 15 and not in ICU” we evaluated the filters’ performance in patients with an ISS greater than 15; for "GCS < 9 and not intubated" we analyzed patients with a GCS less than 9; and lastly, for "No anticoagulants within 72 hours after TBI" we identified patients with head trauma as defined by the precence of at least one head injury according to available AIS codes.
+For each audit filter, we identified the cohort of patients in which the filter was applicable, and evaluated their performance in that cohort: For “ISS > 15 and no team activation” and “ISS > 15 and not in ICU” we evaluated the filters’ performance in patients with an ISS greater than 15; for "GCS < 9 and not intubated" we analysed patients with a GCS less than 9; for "> 30 min until first intervention" and "> 60 min until first CT" we selected patients who underwent a first intervention and those who received a CT respectively.
+Lastly, for "No anticoagulants within 72 hours after TBI" we identified patients with head trauma as defined by the presence of at least one head injury according to available AIS codes.
 
@@
 interventions, level of care and time aspects was retrieved from the trauma registry whereas the exposure, audit filters and the outcome, and opportunities
-for improvement were retrieved from the quality database.
+for improvement were retrieved from the trauma care quality database.
 
@@
 sensitivity, specificity and Cohen's kappa for each audit filter in
-predicting opportunities for improvement. Bootstrapping with 1,000
+predicting opportunities for improvement.
+A Cohen's kappa of 0-0.2 was interpreted as no reliability for the predictions while a value above 0.2 was interpreted as having a minimal reliability up to perfect reliability being 1 [@mchugh2012interrater].
+Bootstrapping with 1,000
 iterations was applied to estimate 95% confidence intervals for all
@@
 according to available AIS codes. Missing values were numbered but
-ignored in the analysis.
+ignored in the analysis for example those without a registered blood pressure was ignored when analysing "SBP < 90".
 
@@
 median age,
-`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` years vs
-`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)` years,
-`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`, median 
-ISS `r inline_text(tableOne, variable = ISS, column = stat_2)` vs
-`r inline_text(tableOne, variable = ISS, column = stat_1)`,
-`r inline_text(tableOne, variable = ISS, column = "p.value")` and
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2, pattern = "{median} (IQR {p25}, {p75})")` years vs
+`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1, pattern = "{median} (IQR {p25}, {p75};")` 
+`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")`) years, median 
+ISS `r inline_text(tableOne, variable = ISS, column = stat_2, pattern = "{median} (IQR {p25}, {p75})")` vs
+`r inline_text(tableOne, variable = ISS, column = stat_1, pattern = "{median} (IQR {p25}, {p75};")`
+`r inline_text(tableOne, variable = ISS, column = "p.value")`) and
 intubations in the emergency department (ED), 
@@
 vs
-`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1)` patients,
-`r inline_text(tableOne, variable = ed_intubated, column = "p.value")`.
+`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_1, pattern = "{N} ({p}%;")`
+`r inline_text(tableOne, variable = ed_intubated, column = "p.value")`) patients.
 Patients with opportunity for improvement had longer times to first major intervention from hospital arrival median
-`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2, pattern = "{median} (IQR {p25}, {p75})")`
 minutes vs
-`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)` minutes,
-`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1, pattern = "{median} (IQR {p25}, {p75};")` 
+`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`) minutes.
 and longer times to first CT, median
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` minutes vs
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)` minutes,
-`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")` minutes.
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2, pattern = "{median} (IQR {p25}, {p75})")` minutes vs
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1, pattern = "{median} (IQR {p25}, {p75};")`
+`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`) minutes.
+
 The frequency of opportunities for improvement was highest in patients
@@
 activation were not significantly correlated with opportunities for
-improvement (p\>0.05). The highest amount of missing data was observed for the audit
-filter "time to first major intervention", which was missing for 5,990 patients, (72.1%). See Table 2
+improvement (p\>0.05).
+The highest amount of missing data was observed with "time to first major intervention", which was missing for 5,990 patients, (72.1%), partly because the absence of an initial CT or major intervention due to the clinical management being observation resulted in their registration as missing.
+See Table 2
 for details of selected patient characteristics.
@@
 The audit filter with the highest sensitivity was "\> 60 min until first
-intervention" (83.3%, 95% CI [79.2-87.5]). The lowest sensitivity was
-seen in "SBP \< 90" (8%, 95% CI [5.6-10.2]) and "Dead at 30 days" (8%,
-95% CI [5.6-10.2]). The audit filter with the highest specificity was
-"Massive transfusion" (95.8%, 95% CI [95.4-96.3]). The lowest
+intervention" (83.3%, 95% CI, 79.2-87.5). The lowest sensitivity was
+seen in "SBP \< 90" (8%, 95% CI, 5.6-10.2) and "Dead at 30 days" (8%,
+95% CI, 5.6-10.2). The audit filter with the highest specificity was
+"Massive transfusion" (95.8%, 95% CI, 95.4-96.3). The lowest
 specificity was seen in "ISS \> 15 and no team activation" (21.6%, 95%
-CI [19.9-23.2]). The sensitivity for all audit filters combined was
-92.6% (95% CI [90.5-95]) and the specificity was 29.4% (95% CI
-[28.4-30.5]).
+CI, 19.9-23.2). The sensitivity for all audit filters combined was
+92.6% (95% CI, 90.5-95) and the specificity was 29.4% (95% CI, 28.4-30.5).
 
 The highest Cohen's kappa was seen for the audit filter "Massive
-transfusion" (0.08, 95% CI [0.05-0.12]). The smallest Cohen's kappa was
-seen in "GCS \< 9 and not intubated" (-0.07, 95% CI [-0.14-0]. The
-Cohen´s kappa for all audit filters combined was 0.04 (95% CI
-[0.03-0.04]. See table 3 for the performance of all audit filters using
+transfusion" (0.08, 95% CI, 0.05-0.12). The smallest Cohen's kappa was
+seen in "GCS \< 9 and not intubated" (-0.07, 95% CI, -0.14-0. The
+Cohen´s kappa for all audit filters combined was 0.04 (95% CI, 0.03-0.04). See table 3 for the performance of all audit filters using
 sensitivity, specificity and Cohen's kappa coefficient.
@@
 
-agree
+Few studies have examined the link between individual audit filters and opportunities for improvement.
+Given the myriad of different audit filters it is challenging to accurately compare our findings with studies using a completely different set of audit filters.
+The delay-related audit filters in this study: "\>60 min until first intervention",
+"\>30 min until first CT" did not predict opportunities for improvement.
+Interestingly, time to definitive treatment and time to first CT was associated with opportunities for improvement in the patient characteristics analysis similar to a previous study by Albaaj et al.
+[@albaaj2023patient] on patient factors associated with opportunities for improvement.
+However, while statistically significant, the observed effect size suggests low clinical relevance.
+Similarly, one study by Lewis et al.
+[@lewis2017delay] found that delays to operating room exceeding 2 hours from emergency department arrival did not individually predict opportunities for improvement.
 
-disagree
-Some previous studies have shown similar findings to this study.
-Lewis et al.
-looked at one audit filter, "delay to operating room" of more than 2 hours and its correlation with opportunities for improvement.
-They found that the audit filter had no significant ability to predict opportunities for improvement.
-Similarly, the delay-related audit filters: "\>60 min until first intervention" and
-"\>30min until first CT" showed noability to predict opportunities for improvement.
-A big meta analysis by Evans et al.
-sought to identify a common set of evidence based audit filters but no study fulfilled the inclusion criteria.
-A 2008 study by Di Bartolomeo et al.
-[@di2008acscot] looked at 4 ASCOT audit filters in relation to 30-day mortality and a morbidity measure.
-One audit filter showed a correlation with mortality but no audit filter showed a correlation with the morbidity measure.
+Furthermore, the audit filter "GCS < 9 and not intubated" showed no correlation with opportunities for improvement in our study.
+Comparably, Willis et al.
+[@willis2008interpreting] found no association between "GCS below 9 and not intubated" and mortality or length of hospital stay but it was associated with shorter stays in the ICU.
+Contrary, Glance et.
+al [@glance2012association] found that "No definitive airway in patient with GCS score ≤ 8" was associated with increased mortality risk but reduced risk of major complication and death.
+This was concluded as reflective of clinical decision rather than quality of care [@willis2008interpreting; @glance2012association].
+For instance, an abnormal GCS because of intoxication
+is managed differently to an abnormal GCS due to head injury.
+Furthermore, Willis et al.
+found that "failure to activate trauma team at major trauma service" was not associated with poor outcomes, similar to our findings.
 
-A recent study by Berg et al. [@berg2024effects], looking at implementation of audit filters on mortality concluded that audit filters might be useful in reducing the mortality. 
-Previous studies suggest poor performance of audit filters for improving
-trauma care [@evans2009audit; @stelfox2010quality; @willis2008interpreting; @di2008acscot; @glance2012association; @lewis2017delay].
-This study adds to that evidence by showing that audit filters perform poorly in predicting opportunities for
-improvement encompassing mortality and non-mortality
-associated outcomes. Some audit filters might be more reflective of clinical decisions rather than quality of care [@glance2012association]. For instance, an abnormal GCS because of intoxication
-is managed differently to an abnormal GCS due to head injury, which could explain the poor performance of "GCS < 9 and not intubated", similarly shown by
-Willis et al. [@willis2008interpreting]. This could also explain the poor performance of the delay-related audit filters: "\>60 min until first intervention" and
-"\>30min until first CT". Similarly, Lewis et al. [@lewis2017delay] showed that delay
-to the operating room was not associated with opportunities for improvement.
-In contrast, Teixeira et al. [@teixeira2007preventable] identified delays to
-treatment and delays to CT as common errors in trauma care. A recent
-study in Japan by Yamamoto et. al [@yamamoto2023immediate] observed
-significantly reduced mortality with whole-body CT scans within 10
-minutes of arrival to the ED.
-If anything, this highlights the unique opportunities for improvement in different hospitals.
-The audit filters also varied greatly in terms of expected mortality.
-Some audit
-filters such as "Massive transfusion" and "Dead at 30 days" have a high
-expected mortality. In such cases the reduced time frame for any potential
-opportunity for improvement to occur could explain their poor performance.
-When audit filters were combined they had a low specificity relative to its sensitivity which corresponds to previous studies showing high numbers of false positives [@sanddal2011analysis; @roy2017learning; @ghorbani2018analysis].
-The sensitivity and specificity for individual audit filters varied substantially, suggesting that some audit filters are too rigid, missing the majority of opportunities for improvement while some are too inclusive, causing high numbers of false positives [@willis2008interpreting].
+Di Bartolomeo et al. [@di2008acscot] looked at the association between four audit filters ("Ambulance scene time >20 min", "Patient with a Glasgow coma
+scale score (GCS) < 13 who does not receive a CT scan of the head within 2 h of arrival at the hospital", "Patients with epidural or subdural haematoma receiving craniotomy >4 h after arrival at ED, excluding those performed for intracranial pressure monitoring", "Abdominal,
+thoracic, vascular or cranial surgery performed >24 h after arrival") and 30-day mortality and 6-month disability. One audit filter: "abdominal, thoracic, vascular or cranial surgery
+performed >24 h after arrival" showed contradictory outcomes with reduced mortality but increased disability.
+This may be explained by aggressive trauma care producing additional survivors with poor functional outcomes or underestimating the mortality rate beyond 30 days.
+The rest of the audit filters showed no significant association with outcomes.
 
-There were some limitations in this study.
-Firstly, it is unlikely that the specific associations presented in this study are generalizable to other hospitals with similar systems, but the broader finding that audit filters are poorly associated with opportunities for improvement is likely to be externally valid [@world2009guidelines].
+A study by Chadbunchachai et al.
+[@chadbunchachai2001comparative] reported a reduction in preventable deaths following the implementation of audit filters, with similar results in a subsequent study [@chadbunchachai2003study].
+In 2009 Evans et al.
+[@evans2009audit] conducted a review to evaluate the effectiveness of audit filters in improving trauma care quality but found no studies of sufficient methodological quality for inclusion.
+The studies by Chadbunchachai et al.
+were excluded due to insufficient pre- and post-implementation data points.
+They concluded that there is no high-quality evidence supporting the use of audit filters in improving trauma care.
+Similarly, Stelfox et al.
+[@stelfox2010quality] found no clearly defined, evidence-based audit filters.
+A more recent study by Berg et al.
+[@berg2024effects] also observed decreased mortality after the implementation of audit filters in India.
+Furthermore, the perceived usefulness of audit filters in trauma quality improvement among health care professionals remains high [@berg2022perceived; @santana2014development].
+
+There are some limitations in this study.
+Firstly, it is unlikely that the specific associations presented in this study are generalizable to other hospitals with similar systems [@di2008acscot; @glance2012association], but the broader finding that audit filters are poorly associated with opportunities for improvement is likely to be externally valid [@world2009guidelines].
 Secondly, instances where the
-same patient case had multiple audit filters introduce uncertainty
+same patient case had multiple audit filters flagged introduce uncertainty
 regarding which specific audit filter truthfully predicted an
-opportunity for improvement. Finally, the number of missing values
+opportunity for improvement, introducing a level of confounding. Finally, the number of missing values
 differed significantly among the audit filters, resulting in varying
-degrees of validity for each filter.
+degrees of validity for each audit filter.
 
@@
 emerging healthcare challenges and resolved quality gaps, reducing their
-utility over time. Therefore, regular evidence-based reviews are needed
+utility over time [@glance2012association]. Therefore, regular evidence-based reviews are needed
 to ensure performance [@santana2014development], though such reviews may
-be cost-prohibitive. Furthermore, they might overshadow other
-contributing factors to opportunities for improvement
-[@willis2008interpreting].
+be cost-prohibitive. Furthermore, they may overshadow other
+contributing factors to opportunities for improvement, i.e confounding variables [@glance2012association; @willis2008interpreting].
 
@@
 successful in predicting opportunities for improvement. Given these
-results, future studies should focus on developing new tools with higher predictive
-ability.
+results, in combination with the lack of high-quality evidence showing that implementing audit filters improve patient outcomes, future studies should focus on developing new tools with higher predictive
+ability to identify opportunities for improvement.
 

### 2024-10-07 main.R

@@ -26,7 +26,7 @@ listOfAuditFiltersClean <- c("SBP < 90", "Dead at 30 days", "ISS > 15 and no tea
 ## viktigt att det ska vara samma ordning ^
 DefOfAbbreviations <- "Definition of abbreviations: SBP = Systolic Blood Pressure; ISS = Injury Severity Score; GCS = Glascow Coma Scale; ICU = Intensive Care Unit; CT = Computer Tomography; ED = Emergency Department; CPR = Cardiopulmonary Resuscitation; TBI = Traumatic Brain Injury"
 
-selectedAuditFilter <- listOfAuditFilters[1:11]
+selectedAuditFilter <- listOfAuditFilters[1]
 
 tableKappa <- data.frame(Auditfilter = character(0),
                          Sensitivity = numeric(0),
